---
editor_options: 
  markdown: 
    wrap: 72
---

# Lower Extremity

**UNDER CONSTRUCTION - This Chapter is Currently in DRAFT mode**

This chapter is lengthy and covers all vascular disease of the lower
extremity, progressing in sub-chapters through peripheral arterial
occlusive disease, critical limb ischemia, acute limb ischemia,
compartment syndrome, lower limb amputation, and some of the rarer
non-atheromatous conditions.

## Pathophysiology of Peripheral Arterial Occlusive Disease

**Accompanying Episode:** Peripheral Arterial Occlusive Disease (Part
1): Nedal Katib and Danielle Bajakian

```{r echo=FALSE}

library(embedr)

embed_audio("https://traffic.libsyn.com/audiblebleeding/PAOD_part_1_postEQ.mp3")

```

**What is Peripheral Artery Occlusive Disease (PAOD)?**

Peripheral Arterial Disease encompasses extremity arterial disease but
generally is used to describe lower limb arterial occlusive disease.
Peripheral Artery Occlusive Disease (PAOD) is a more specific term and
this encompasses atherosclerotic disease of the lower limb arteries.

The disease has a step-wise spectrum of presentations from asymptomatic
disease to intermittent claudication and finally chronic limb
threatening ischemia (CLTI) formerly known as Critical Limb Ischemia
(CLI) [@aboyans2017ESCGuidelines2018]

**What is the underlying pathophysiology of PAOD?**

The understanding of the underlying pathophysiology which results in
occlusive arterial disease of the lower limb has evolved over the last
50 years. Atherosclerosis remains the main pathological process
resulting in occlusive disease with smoking being the primary modifiable
risk factor.\
Although with the rising incidence of diabetes a second population of
patients with a significant PAOD burden has developed.

Atherosclerosis begins with an injury to the intimal lining of the
arterial wall, which, for example, can result from smoking,
hypertension, hyperglycemia, or advanced age. Repeated injury leads to a
chronic inflammatory process resulting in the uptake and oxidation of
LDL, resulting in foam cells.[@hiltunen1998; @tsimikas2005;
@witztum1994] Foam cells result when unregulated scavenger receptors
promote uptake of oxidized LDL by marophages and smooth muscle cells.
These foam cells become necrotic and then create the central core of the
atheroma.[@owens2019] This oxidized LDL form into intimal plaque, build
up and calcification that may result in progressive vessel stenosis and
subsequent occlusion, or plaque rupture with acute occlusion.

Atherosclerosis is particularly prone to developing at sites of arterial
bifurcation (i.e. carotid bifurcation, femoral bifurcation) because
turbulent flow and low laminar shear stress results in the down
regulation of eNOS and NO production. This in turn increases V-CAM-1 and
reduces inhibition of KF-kB (resulting in it's elevation.[@owens2019]

| Classification | Description                                                             |
|:---------------:|-------------------------------------------------------|
|       0        | Normal artery                                                           |
|       1        | Initial atherosclerotic lesion                                          |
|       2        | Fatty streak                                                            |
|       3        | Fatty plaque/raised fatty streak                                        |
|       4        | Atheroma                                                                |
|       5        | Fibroatheroma - development of fibrotic cap, vascularization or calcium |
|       6        | Complicated lesion                                                      |

: AHA lesion classification - defines plaque risk for
rupture/thrombosis. Extrapolated from the coronary artery disease
literature.[@mcgill2000]

**What are the risk factors for atherosclerosis?**

**Modifiable:**

1.  ***Smoking***

    -   The **most significant modifiable risk factor** for developing
        peripheral arterial disease.

    -   Causes endothelial dysfunction by reducing nitric oxide and
        triggering reactive-oxygen species, platelet ahesion, and
        permeability of endothelial surface to
        fibrinogen.[@unitedstatessurgeongeneralHealthConsequencesSmoking2014;
        @hackam2003].

    -   Causes a prothrombotic environment by causing an increase in
        thromboxane A2 and decreasing prostacyclin thus overall
        resulting in increased prothrombotic environment for platelets.

    -   Smoking has a stronger association with Intermittent
        claudication than with Coronary Artery Disease!
        [@gordonPredispositionAtherosclerosisHead1972]

2.  ***Diabetes, Metabolic Syndrome, and Insulin Resistance***

    -   Diabetes Mellitus, after smoking, is the most significant
        modifiable risk factor for developing peripheral arterial
        disease. Both insulin resistance and hyperinsulinemia are
        independent risk factors for developing peripheral arterial
        disease.

    -   The Odds Risk for developing PAD in patients with DM ranges from
        1.89 to 4.05.

    -   Factors associated with developing PAD in diabetics include
        female gender, chronicity/severity of DM, insulin dependence,
        African American race or hispanic ethnicity.[@regensteiner2015;
        @selvin2006; @wattanakit2005] An increase in HbA1C by 1%
        correlates with a 28% increase risk of developing PAD.
        [@adlerUKPDS59Hyperglycemia2002]

    -   The following mechanisms may lead to PAD in diabetics

        -   Hyperglycemia reduces bioavailability of NO, which is a
            vasodilator, enhances endothelial function, and inhibits
            both platelet activation and smooth muscle
            migration.[@higashi2009]

        -   Hyperglycemia activates protein kinase C and thus
            accumulation of mitochondrial ROS and
            prostanoids.[@geraldes2010]

        -   Endothelial dysfunction leads to increased endothelin-1 and
            vasoconstriction.[@ahlborg2007]

3.  ***Hypertension***

    -   The **most common** cardiovascular risk factor worldwide.

        -   USPSTF Grade A recommendation for HTN screening confirm
            readings outside clinical setting or at home before
            initiating treatment.[@piper2014]

    -   The Incidence of PAD increases to 2.5-fold in patients with
        Hypertension. [@kannelUpdateEpidemiologicFeatures1985]

    -   Each 20mg increase in BP, results in a 35-63% increase risk in
        PAOD. Recommend keeping BP \<140/90 or \<130/80 in diabetics or
        CKD.[@adler2000; @bavry2010; @emdin2015; @itoga2018]

4.  ***Dyslipidemia***

    -   A strong association has long been identified as a risk factor
        for cardiovascular disease.[@langlois2018; @diehm2004]

    -   25% cardiovascular event reduction for each 39 mg/dL (1mmol/L)
        reduction in LDL.
        [@heartprotectionstudycollaborativegroupRandomizedTrialEffects2007]

    -   JUPITER Study - Randomized rosuvastatin 20mg to placebo in
        healthy patients. Found significant decrease in unstable angina,
        myocardial infarction and stroke. However, patients had a higher
        incidence of physician-reported diabetes, but no difference in
        myopathy.[@ridker2008]

    -   Fish oil is also indicated for secondary prevention among
        patients with known cardiovascular disease.[@grenon2012;
        @kris-etherton2002; @siscovick2017]

**Non-Modifiable:**

1.  ***Age***

    -   Age is identified as a risk factor for PAD regardless of gender.

    -   Prevalence of PAD increases with age: 15% \> 70 years of age.

2.  ***Gender***

    -   The Framingham Study has found that the risk of developing PAD
        is doubled in men.

3.  ***Race and Ethnicity***

    -   The MESA study showed a higher prevalence of PAD (ABPI \<0.9) in
        African Americans compared to Whites. 7.2% versus 3.6%.
        [@bildMultiEthnicStudyAtherosclerosis2002]

    -   Cross Sectional analysis 6653 subjects all with ABPI assessment
        revealed a prevalence of PAD (\<0.9) of 4%. Non-Hispanic Whites:
        3.6%, Asian: 2%, African American: 7.2% and Hispanic: 2.4%.
        (p\<0.01) [@allisonEffectNovelCardiovascular2006]

**What are the some of the major population-based trials looking at the
natural history of atherosclerotic disease?**

1.  ***The Framingham Heart Study***: The original Cohort from the town
    of Framingham, n=5183 patients, followed over time for over 30
    years. There have been multiple subsequent recruited populations
    since. The majority of information we have about risk factors
    related to cardiovascular health comes from this study.
    [@mahmoodFraminghamHeartStudy2014]

2.  ***The Rotterdam Study***: 1990, Longitudinal Study, \>7000
    participants.

3.  ***CVHS***: 1989-1999 Longitudinal Study : n\>5000 Multicenter
    Study.

4.  ***MESA***: Cross Sectional analysis 6653 subjects all with ABPI
    assessment revealed a prevalence of PAD (\<0.9) of 4%. Non-Hispanic
    Whites: 3.6%, Asian: 2%, African American: 7.2% and Hispanic: 2.4%.
    (p\<0.01) [@allisonEffectNovelCardiovascular2006]

5.  ***The Edinburgh Study***: The EAS began as a cross sectional study
    of 1592 men and women in Edinburgh with the goal of examining the
    frequency of risk factors for peripheral arterial disease. The
    subjects were followed over 20 years.
    [@fowkesEdinburghArteryStudy1991]

**How does Diabetes confound the clinical picture of PAOD?**

-   Increasing Incidence of Diabetes
    world-wide.[@boultonGlobalBurdenDiabetic2005]

    -   2.8% of the global population affected by diabetes in 2000,
        estimated to be 4.4% by 2030.

    -   25% of patients with diabetes develop a diabetic foot ulcer
        (DFU), at some stage in their lives.

    -   Diabetic foot ulcers have a very high rate of recurrence with
        40% of patients developing a new ulcer within 12 months of
        healing a previous ulcer. [@armstrong2017]

    -   Limb Loss occurs every 20 seconds world-wide to due to Diabetes
        related PAOD or infection.

**What's the pathophysiology of Diabetes and PAOD**
[@armstrong26PathophysiologyPrinciplesManagement2012]

-   Neuropathy. The loss of the basic nociceptive mechanisms in the foot
    among diabetics, presents as a loss of protective sensation (LOPS).
    Neuropathy can be divided into three types:

    -   Sensory: "stocking-glove" distribution

    -   Motor: Intrinsic muscle wasting -- resulting in deformities

    -   Autonomic: Sympathetic nervous system pathology

-   Structural deformaties and gait distrubances

-   Arterial disease of large and small vessels.

    -   Large Vessel

    -   Small Vessel

    -   Both

-   Foot infection[@bandyk2018; @kayssi2019; @lepäntalo2011]

Given the pathological effect of diabetes upon multiple structures and
systems the normal progressive history of diabetes related PAOD is
somewhat different to smoking related PAOD. Clinically due to the
combined effect of peripheral neuropathy and microvascular arterial
disease, foot ulceration for many patients tends to be the initial
presentation. This results in a lack of a 'safety net' where presenting
with progressive claudication allows for a period of detection,
management and and risk factor modification before they develop tissue
loss and are at risk of amputation, leading to poor outcomes and greater
morbidity and mortality.

Since the publication of WIfI in 2014 a lot has changed in the way we
view PAOD leading up to last years new Global Vascular Guidelines on
CLTI, which as a term has replaced CLI (see
\@ref(chronic-limb-threatening-ischemia)).
[@millsSocietyVascularSurgery2014a; @conteGlobalVascularGuidelines2019a]

## Intermittent Claudication

**What is Intermittent Claudication and the classic patient
presentation?**

The original population studies we mentioned determined the epidemiology
and natural history of Intermittent Claudication based on historically
validated and widely accepted questionnaires, namely the ***Rose***
[@roseDiagnosisIschaemicHeart1962] (which later was adopted by the WHO)
and subsequently the ***Edinburgh questionnaire***.
[@lendEdinburghClaudicationQuestionnaire1992].

All questionnaires are based on a number of key diagnostic clinical
factors that define claudication, they are:

-   Onset

-   Calf Involvement

-   Reproducibility

-   Relief with Rest

-   Not Occurring at Rest

The progression of intermittent claudication was historically graded by
Fontaine (1954) [@fontaineSurgicalTreatmentPeripheral1954] followed by
the Rutherford Grading System (1986, Revised 1997)
[@rutherfordRecommendedStandardsReports1997]

Rutherford et al. Ad Hoc Committee on Reporting Standards, SVS/North
American Chapter ISCVS:

| Grade | Category |                     Clinical Description                     |
|:----------------------:|:----------------------:|:----------------------:|
|   0   |    0     | A symptomatic - no hemodynamic significant occlusive disease |
|   I   |    1     |                      Mild Claudication                       |
|   I   |    2     |                    Moderate Claudication                     |
|   I   |    3     |                     Severe Claudication                      |
|  II   |    4     |                      Ischemic Rest Pain                      |
|  III  |    5     |                      Minor Tissue Loss                       |
|  III  |    6     |                      Major Tissue Loss                       |

**Leriche Syndrome** - Is a clinical triad of: claudication (buttock,
thigh, and calf), impotence, and decreased lower limb pulses, which
signifies aortoiliac occlusive disease. [@frederick2010; @leriche1948;
@setacci2012]

**Neurogenic Claudication** - Differentiating neurogenic claudication
from vasculogenic claudication can be difficult. Vasculogenic
claudication is relieved by cessation of ambulation and mostly affects a
unilateral calf. Neurogenic improves with postural changes, more often
bilateral and affects the thighs.
[@nadeauReliabilityDifferentiatingNeurogenic2013]

**What is involved in the work up of patients with PAOD/Intermittent
Claudication?**

### History / Clinical Examination

***SVS Guidelines:***

"We recommend using ABPI as the first-line non-invasive test to
establish a diagnosis of PAOD in individuals with symptoms or signs
suggestive of disease. When the ABI is borderline or normal (\>0.9) and
symptoms of claudication are suggestive, we recommend an exercise ABPI."

*Grade 1 Level of Evidence A*

-   ABPI

-   Exercise ABPI

-   Ultrasound

**What is an ABPI and how is it measured?**

The AHA released guidelines on how to perform an ABPI in an attempt to
standardize the method to allow for more comparable results from
studies:

Divide the higher of the PT or DP pressure by the higher of the right or
left Brachial SBP *(Class 1 Level of Evidence A)*
[@aboyansMeasurementInterpretationAnkleBrachial2012]

Sensitivity and Specificity both \>95% (when ABPI cut off \</=0.9 -- in
detecting \>/= 50% stenosis) [@yaoAnkleSystolicPressure2005;
@ourielCriticalEvaluationStress1982]

|  **ABPI**  |  Interpretation  |
|:----------:|:----------------:|
|   \>1.4    | Non-compressible |
| \>0.9-1.39 |      Normal      |
|  0.5-0.9   | Mild to Mod PAD  |
|  0.0-0.5   |    Severe PAD    |

: Interpretation of ABPIs.

**What are Exercise ABPI studies?**

Constant Load Testing -- (unlike the Graded Test -- Bruce Protocol).

The physiological changes in ABPI related to exercise are due to the
fact that exercise increases SBP and decreases SVR and, therefore, will
affect flow through a stenosis making it more physiologically relevant
during exercise. [@nicolaiReliabilityTreadmillTesting2009;
@alqahtaniPredictorsChangeAnkleBrachial2018]

Walking distance has been shown to correlate with level and severity of
POAD. [@strandnessHemodynamicsSurgeons1975a]

Furthermore patients with single level iliac disease may present with
buttock/thigh claudication with preserved pulses. Exercise treadmill
ABPI particularly useful in determining the severity of disease in these
patients. [@aboyansGeneralPrognosisPatients2010]

For more on exercise testing see \@ref(lower-extremity).

**What is the ultrasound duplex criteria for defining PAOD?**

| Stenosis Category |       Peak Systolic Velocity       | Velocity Ratio | Distal Artery Spectral Waveform |
|:----------------:|:----------------:|:----------------:|:----------------:|
|      Normal       |               \<150                |     \<1.5      |      Triphasic, Normal PSV      |
|      30-49%       |              150-200               |     1.5-2      |      Triphasic, Normal PSV      |
|      50-75%       |              200-400               |      2-4       |     Monophasic, reduced PSV     |
|       \>75%       |               \>400                |      \>4       | Damped, monophasic, reduced PSV |
|     Occlusion     | No Flow -- B -mode, Terminal Thump |       NA       |               NA                |

Adapted from Stone and Hass. Vascular Laboratory: Arterial Duplex
Scanning. Rutherford's Vascular Surgery and Endovascular Therapy.
2019.[@stone21VascularLaboratory2019]

**What Guidelines are there pertaining to PAOD Management.**

***SVS Guidelines (2015)***

-   The SVS published the SVS practice guidelines for atherosclerotic
    occlusive disease of the lower extremities:**Management of
    asymptomatic disease and claudication**. *Conte and Pomposelli et
    al. JVS 2015* [@conteSocietyVascularSurgery2015]

Other Guidelines:

-   TASC 1 -2

-   European Guidelines (2017)

-   AHA Guidelines (last update 2016)

### Management

#### Medical Management

**What is the initial management of Asymptomatic Patients with PAOD?**

1.  Smoking Cessation -- Multidisciplinary comprehensive smoking
    cessation interventions -- repeatedly until tobacco use has stopped
    (Grade A -- 1)
2.  Intervention is not only not recommended, but invasive treatment is
    recommended against, in the absence of symptoms (Grade A -1)

**How can we medically manage asymptomatic PAOD based on the SVS
Guidelines?**

-   ***Anti-Platelet Therapy***

    -   The Aspirin for Asymptomatic Atherosclerosis Trial -- n=3350,
        aspirin versus placebo. 8 years follow up no difference in
        events [@fowkesAspirinPreventionCardiovascular2010] -- therefore
        benefit unknown
    -   Literature reviews find benefits in secondary prevention with
        antiplatelet monotherapy (aspirin or clopidogrel) in symptomatic
        PAD.[@banerjee2015] Some initial studies did demonstrate benefit
        from vorapaxar as well.@rooke2013

-   ***Statin Therapy***

    -   The Heart Protection Study
        [@heartprotectionstudycollaborativegroupRandomizedTrialEffects2007] -
        this study looked at statins in patients with PAD but not
        completely asymptomatic, they had other risk factors such as
        diabetes, IHD, cerebrovascular disease, or hypertension. Without
        these risk factors the benefit of Statin therapy remains
        unclear.

    -   However, the AHA from the Framingham Study does recommend using
        statins if 10-year risk based on risk calculators \>7.5% (which
        would be positive if PAOD present).

    -   Statins are particularly useful for risk reduction of
        cardiovascular events in
        diabetics.[@cholesteroltreatmenttrialists(ctt)collaborators2008;
        @colhoun2004; @stamler1993]

    -   Statin-associated muscle symptoms (SAMS) is the most common side
        effect and occurence has been reported as high as 30% in some
        cohorts.[@keen2014] Other side effects are rare (1.5-5% of
        patients) and include DM2 new onset, neurocognitive effects,
        hepatotoxicity, renal toxicity and pancreatitis.[@bitzur2013;
        @saxon2016; @ward2019]

-   ***Exercise and Limb Function***

    -   No clear evidence that physical therapy improves the patient
        outcomes or quality of life.

-   ***Surveillance***

    -   No benefit from US surveillance, unclear benefit of ABPI
        surveillance.
    -   Patients are at higher risk of depression (almost 20%),
        therefore PHQ-9 and -2 are good screening tools.[@jha2019;
        @mcdermott2016; @ramirez2018]

**How can we medically manage Intermittent Claudication based on the SVS
Guidelines?**

-   **Smoking Cessation:** Multidisciplinary comprehensive smoking
    cessation interventions -- repeatedly until tobacco use has stopped
    *(Grade A -- 1)* First line therapy is behavioral therapy +
    varenicline/bupropion or nicotine replacement therapy (NRT) per
    EAGLES trial.[@anthenelli2016; @vogeler2016]

    -   Preloading with varenicline (4wks) or bupropion (5-7d) prior to
        quitting has demonstrated higher abstinence at
        12w.[@ratchford2011]

        -   Varenicline is a partial nicotine receptor
            agonist.[@cahill2016]

        -   Buproprion is a selective dopamine and norepinephrine uptake
            inhibitor.[@hughes2014]

    -   NRT is preferred if rapid onset of action is
        needed.[@thomsen2014] However, contraindicated in setting of
        recent MI, severe angina, or life threatening arrhythmias.

-   **Dyslipidaemia:** Statin Therapy Recommended -- most recent
    evidence on lipid therapy has suggested focusing on reducing 10-year
    cardiovascular event risk rather than specifically reducing lipid
    levels. *(Grade 1-A)*

-   **Anti-Platelet Therapy:** Aspirin therapy (75-325mg daily) is
    recommended to reduce cardiovascular events in patients with PAD
    *(Grade 1-Level A)*
    [@collaborationCollaborativeMetaanalysisRandomised2002;
    @antithrombotictrialists(att)collaboration2009]

    -   There is evidence that Clopidogrel 75mg compared to Aspirin is
        better in event reduction.
        (CAPRIE)[@capriesteeringcommitteeRandomisedBlindedTrial1996]--
        replacing Aspirin with Clopidogrel *(Grade 1-Level B)*
    -   Recent evidence from the COMPASS trial has suggested that the
        addition of low dose rivaroxaban (2.5mg BID) to aspirin (100mg
        daily) has lowere MALE, composite endpoint of cardiovascular
        death, stroke and MI. Subgroup analysis suggested clopidogrel
        may be better with reduced risk of GI bleed and
        MACE.[@anand2018; @eikelboom2017]

-   **Diabetes Mellitus:** -- Optimization of HbA1C \< 7% *(Grade
    1-Level B)*[@lajoie2019]

-   **Hypertension** -- Indicated B-Blockers for hypertension. *(Grade
    1-Level B)* (there's no evidence that Beta Blockers worsens IC)

    -   Initiation of beta-blockers in the immediate preoperative period
        is associated with worse outcomes.[@blessberger2018;
        @poisestudygroup2008]

    -   ACE Inhibitors are first line in patients with heart failure,
        diabetes or kidney disease but should be avoided in patients
        with renal artery stenosis (RAS).[@barrons2016]

-   **Homocysteine** -- Recommendation against Folic Acid and Vit
    B12*(Grade 2 -- C)*

    -   Higher homocysteine levels are associated with PAOD and risk of
        venous and arterial thrombosis.[@dangelo1997; @khandanpour2009;
        @liebman2019]

**To improve Limb Function in patients with IC:**

-   **Cilostozol** use -- IC without CHF -- 3-month Trial (Grade 2 - A)

    -   If unable to tolerate Cilostozol -- Pentoxifylline (400mg TDS)
        *(Grade 2 -- B)*

    -   Based on Meta-analysis 26 trials
        [@stevensSystematicReviewEfficacy2012]

-   **Supervised Exercise Therapy**

    -   First Line Therapy recommended SEP: minimum three times / week
        (30-60 min/session) for at least 12 weeks (*Grade 1 Level A*)

    -   More durable functional improvements than angioplasty alone,
        however no current data demonstrating impact on need for
        revascularization or amputation rate.[@fakhry2013; @fakhry2015;
        @murphy2012; @murphy2015]

    -   Meta-analysis of 32 RCTs: Placebo versus exercise: Walking Time,
        Walking Ability, Pain Free Walking and maximum walking distance
        improves. BUT no difference in ABPI, Mortality or amputation.
        [@laneExerciseIntermittentClaudication2017]

    -   Meta-analysis of 14 RCTs: SEP better than Non-Supervised
        Programs. [@hagemanSupervisedExerciseTherapy2018]

#### Open and Endovascular Management

**What is the surgical management for patients with Intermittent
Claudication?**

-   With appropriate medical management and risk modification, 25% of
    patients will improve, 50% will remain stable and 25% will progress
    and requiring intervention. [@conte2015; @norgren2007]

-   **Patient Selection for Intervention:**

    -   20-30% of patients with IC who adhere to risk factor
        modification will have progressive symptoms that will eventually
        be treated with intervention.

    -   Patient selection should be based on QoL and functional
        impairment in an active person (loss of ability to perform
        occupation or that limits basic activities of daily living)
        rather than hemodynamic (ABPI or US) or anatomical disease
        progression/severity.

-   **Always remember multifactorial causes of immobility --
    particularly in the elderly.**

    -   SVS recommends that invasive therapy for IC have a \>50%
        likelihood of sustained clinical improvement for at least 2
        years.

-   **Anatomical Selection:**

    -   ***Aortoiliac Disease:***

        -   Previous TASC Classification has attempted to categorize
            anatomy of disease and subsequent recommendation of
            Endovascular versus open surgery. But as the authors of the
            SVS guidelines highlight, "improvements in technology and
            endovascular techniques have resulted in EVT replacing open
            surgical bypass as a primary treatment for both focal and
            advanced AIOD in many cases."

        -   **The majority of evidence is non randomized and meta
            analyses of non-randomized series.**

        -   Endovascular procedures over open surgery for focal AIOD
            causing IC. *(Grade 1 Evidence B)*

        -   Endovascular interventions as first line for CIA or EIZ
            occlusive disease-causing IC. *(Grade 1 Level B)*

        -   Hybrid recommended for Iliac disease involving CFA. *(Grade
            1 Level B)*

        -   Direct Surgical reconstruction (bypass, endarterectomy) in
            patients with reasonable surgical risk and diffuse AIOD not
            amenable to endovascular approach, after one or more failed
            attempts at EVT, or combined occlusive and aneurysmal
            disease. *(Grade 1 Evidence B)*

    -   ***Infrainguinal Disease:***

        -   Reviewing historical data comparing all EVT compared with
            surgical bypass EVTs are less durable, especially when
            there's diffuse or long segments of occlusion/multilevel
            infrainguinal disease.

        -   **Most recommendations are based on low level evidence when
            comparing EVT versus Open Surgery**

        -   Focal + Not involving SFA origin = EVT *(Grade 1 Level C)*

        -   SFA 5-15cm, self-expanding stent (with or without
            paclitaxel) *(Grade 1 Level B)* -- NB: (This was in 2015 pre
            Katsanos Paper[@katsanos2018])

            -   After SFA stenting, SVS and ACC recommend at least 30d
                dual anti-platelet, then single anti-platelet or
                anti-platelet and rivaroxaban.[@hussain2018;
                @societyforvascularsurgerylowerextremityguidelineswritinggroup2015;
                @strobl2013; @tepe2012]

        -   Recommend against infrapopliteal treatment for IC *(Grade 1
            Level C)*

        -   Initial Surgical Bypass *(with vein: Grade 1 Level A)*: If

            -   Diffuse FP Disease

            -   Small Caliber \<5mm

            -   Extensive Calcification in SFA

            -   Average or Low Operative Risk *(Grade 1 Level B)*

## Chronic Limb Threatening Ischemia

**Accompanying Episode:** PAOD (Part 2), Chronic Limb Threatening
Ischemia: Nedal Katib and Danielle Bajakian

```{r echo=FALSE}

library(embedr)

embed_audio("https://traffic.libsyn.com/audiblebleeding/PAOD_part2_postAuphonic.mp3")

```

### Etiology and Presentation

**How does PAOD pathologically progress clinically into Chronic Limb
Threatening Ischemia (CLTI)?**

The latter stages of both the Rutherford and Fontaine Classification
systems highlight this progression, with the Rutherford classification
of Stage 5 being specifically minor tissue loss with focal gangrene, and
stage 6 as major tissue loss identified by spreading of gangrene beyond
the Trans metatarsal level.

Rutherford et al. Ad Hoc Committee on Reporting Standards, SVS/North
American Chapter ISCVS:

| Grade | Category |                    Clinical Description                    |
|:----------------------:|:----------------------:|:----------------------:|
|   0   |    0     | Asymptomatic -no hemodynamic significant occlusive disease |
|   I   |    1     |                     Mild Claudication                      |
|   I   |    2     |                   Moderate Claudication                    |
|   I   |    3     |                    Severe Claudication                     |
|  II   |    4     |                     Ischemic Rest Pain                     |
|  III  |    5     |                     Minor Tissue Loss                      |
|  III  |    6     |                     Major Tissue Loss                      |

**What is Chronic Limb Threatening Ischemia (CLTI), sometimes previously
called Critical Limb Ischemia (CLI)?**

In the last decade leading up to the 2021 Global Vascular Guidelines
(GVG), the term (Chronic Limb threatening Ischemia) CLTI has been
gradually replacing CLI. The GVG mentions that their "promotion" of the
term CLTI is partly due to terms such as "critical or severe limb
ischemia" failing to "recognize the full spectrum and inter-relatedness
of components beyond ischemia that contribute to major limb
amputation..."

**What was the original definition and threshold for CLI, and how can we
make sure we elicit the right symptoms from the patient?**

John Cranley back in his publication in 1969 defined Ischemic Rest Pain
as,

"...pain that occurs in the toes or in the area of the metatarsal heads.
Occasionally...in the foot proximal to the metatarsal heads. Elevation
of the limb above or at the horizontal position aggravates the pain and
pendency...brings relief..." [@cranleyIschemicRestPain1969]

***Nocturnal Rest Pain:*** Worse due to horizontal positioning (no
gravitational assistance) and systolic BP drop during sleep.

### Evaluation

**What aspects of the clinical assessment is important?**

Clinical Assessment involves a full history (the differential mentioned
in Intermittent Claudication) and examination.

-   **Clinical Examination:**

    -   ***Buergers Test (Beurger 1908) / AKA Ratschows Test***
        (Max-Ratschow-Klinic) identifies when there is critical ischemia
        without necrosis yet or gangrene, and is characterized by pallor
        when the leg is elevated above the level of the heart, which
        then turns red when hanging down over the edge of the bed. This
        redness is referred to as "Sunset appearance" and its due to
        abnormal autoregulation. Its been described that normally only a
        third of the capillary bed is open at any time but in a state of
        critical ischemia because of the autoregulation being paralyzed
        a significantly higher portion of the capillaries open up.

    -   ***The ischemic Angle:*** A refinement to Berger's Test: The
        angle of elevation from the horizontal at which the Doppler
        Signal of the PT or DP disappears. This is also referred to as
        the 'pole test', whereby the foot is raised alongside a
        calibrated pole marked in mmHg.

-   **Tissue Loss:**

    -   Gangrene Dry or Wet (infection)

    -   Level of tissue Loss

    -   Probing To Bone/ Exposed structures: Tendons, Soft Tissue, bone,
        Joint Capsule.

    -   Examination of an Ulcer (may have many etiologies) -- important
        not only to identify extent of disease but also to exclude other
        etiologies:

        -   Such as venous, mixed, infective, autoimmune, inflammatory,
            malignancy or trauma.

-   **Foot Infection** - Signs of infection, erythema, rubor,
    cellulitis, tenderness or unexplained hyperglycemia in diabetic
    patients should prompt urgent referral.[@kalish2010]

**What is the natural history of CLTI and what do we know about its
prognosis?**

Fortunately, only a small portion of patients with Intermittent
Claudication will go on to develop rest pain or tissue loss. Its
estimated that anywhere between 5% -29% of patients with PAD or IC go on
to develop CLTI over 5 years.

However, those that do develop CLTI, have a high risk of limb loss
(greater than 20% annual risk).

CLTI measurements most predictive of non-healing are ankle pressure
\<50mmHg, ABI \<0.4, TcPO2 \<20mmHg, and TP
\<20mmHg.[@gerhard-herman2017; @wickström2017]

Patients with CLTI have a high mortality risk (10-15% annual risk), the
majority of terminal events being related to cardiovascular events. Limb
loss or mortality may reach as high as 50% in 1 year. [@adam2005;
@norgren2007]

**What has changed in the last few decades?**

Etologically the prevalence of smokers(Ex and Current) in the population
has decreased and the prevalence of diabetes has increased.

**What is the WIFI Classification?** [@millsSocietyVascularSurgery2014a]

Interestingly in the original article by Bob Rutherford regarding
Diabetes and PAOD:

\- "It was generally agreed that diabetic patients who have a varied
clinical picture of neuropathy, ischemia and sepsis make the definition
even more difficult and it is desirable that these patients be
excluded...diabetic patients should be clearly defined as a separate
category or should be clearly defined as a separate category."

Since then, the SVS, while acknowledging that we can no longer exclude
these patients and treat them separately given the overlap, have decided
that a new classification system is necessary, as one of the key authors
(Joseph Mills) states:

-   "We classify things into groups to differentiate, remember and
    compare, observe and predict their behavior over time." --Joseph
    Mills

***WIfI stands for: Wound, Ischemia and foot Infection***. Most of the
existing Vascular and non-Vascular classification systems don't include
all three components or fail to stratify the degree of ischemia and
presence of gangrene.

Principles of WIfI:

1.  ***Grades, Classes and Stages*** -- Each of the three categories
    (WIfI) have Grades 0,1,2,3: Resulting in 64 Classes.

2.  ***Delphi Consensus*** -- Clinical Stages 1 (Very Low),2 (Low), 3
    (Moderate), 4 (High Risk/Benefit).

    1.  What is the one-year risk of amputation with medical therapy
        alone?

    2.  What is the potential benefit from successful revascularization?

    3.  Analogous to TNM Staging

"It is intended to be an iterative process with the goal of more
precisely stratifying patients according to their initial disease
burden, analogous to TNM cancer staging, but not to dictate therapy."

**What about a differential diagnosis or other causes of similar pain as
rest pain?**

Acute lower limb ischemia has a different clinical presentation, but
there may some overlap with Acute on Chronic disease such as in the case
of in situ thrombosis in the lower limb arterial system.

Other causes of ischemic pain include:

-   Buergers Disease, or Thromboangiitis Obliterans - for more see
    \@ref(buergers-disease)

-   Scleroderma

-   Fibromuscular Dysplasia

-   Popliteal Artery Entrapment

-   Cystic Adventitial Disease

-   Persistent Sciatic Artery Disease

**What is the Rutherford Acute Ischemia Grading System?**

Although acute ischemia is very different from chronic ischemia,
patients with progressive chronic PAD can develop an acute picture
whether from embolism or in-situ thrombosis secondary to plaque rupture.
See \@ref(acute-limb-ischemia) for more.

**What are the current CLTI Guidelines?**

In 2019 the SVS, the ESVS and the World federation of Vascular Societies
(WFVS) joined forces to put together the structure and funding of the
***Global Vascular Guidelines Initiative (GVG)***. Importantly all
sponsorship was directly from the societies and any direct industry
sponsorship or external sources were excluded. They put together a
steering committee responsible for recruiting a large and diverse
writing group and outlined the scope and developed the section briefs of
the guideline.

They determined that:

-   "The term"critical limb ischemia" (CLI) is outdated and fails to
    encompass the full spectrum of patients who are evaluated and
    treated for limb-threatening ischemia in modern practice."

-   CLTI was promoted as the term of choice and was defined by the
    target population.

-   The target population were:

    1.  ***Ischemic Rest Pain*** with confirmatory hemodynamic studies.

    2.  ***Diabetic Foot Ulcer*** or any lower limb ulceration present
        for at least 2 weeks.

    3.  ***Gangrene*** involving any portion of the lower limb or foot.

-   Exclusion from the population:

    1.  ***Purely Venous Ulcers***

    2.  ***Acute Limb Ischemia***/acute trash foot/ischemia due to
        emboli

    3.  ***Acute Trauma*** or mangled extremity

    4.  ***Wounds secondary to non-atherosclerotic conditions***

Methodology of the guidelines utilized the structure of the ***GRADE***
certainty of evidence system.

They highlighted particular important sections in the evaluation and
management of patients with CLTI: ***Patient Risk stratification***,
***Limb Assessment and Severity of Limb Threat*** and the development of
a specific ***evidence-based revascularisation guideline in CLTI***.

It is important to note that compared to most guidelines, unfortunately
in the management of CLTI, particularly when it comes to
revascularisation, the level of evidence is generally **LOW**. Again, it
is important to highlight that the significance of these guidelines in
developing a standard approach and appropriately stratifying patients in
not only management but ongoing research.

**What clinical evaluation is necessary for the patient with CLTI?**

In addition to history, examination, and WIfI assessment mentioned
above, for patients with diabetes and an ulcer a full assessment of
neuropathy and a "probe to bone" test for any open ulcers is recommended
as part of good practice.

In the PAOD and intermittent claudication sub chapters the non-invasive
methods of assessment for these patients have been discussed. In
addition it is important to emphasize the role of Toe Pressures (TP) and
Toe Pressure Index (TPI) in this cohort of patients.

Healing of an ulcer or tissue loss is unlikely if a patient's toe
pressures are less than 55mmHg. And Toe Pressures have been validated in
multiple studies to correlate with Amputation free survival and wound
healing: Amputation Free Survival TP \<30mmHg 2.13 HR (1.52-2.98).
[@wickström2017; @hicks2018]

Patients with ESRD or DM develop medial calcification and often have
elevated ABPI (\>1.3) - which is associated with an elevated risk of
cardiovascular mortality. Outcomes in patients with ESRD are worse in
relation to amputation free survival and amputation rates, regardless of
revascularization strategy.@meyer2018 Toe pressures are particularly
important in this scenario. [@resnick2004; @vitti1994]

Regarding non-invasive assessment for wound healing, a tcPO2 greater
than 40mmHg has the greatest correlation with amputation stump
healing.[@malone1987]

### Management

**What are the recommendations for patients with CLTI when it comes to
Medical Therapy and Risk Factor Modification?**

-   Treat all patients with CLTI with an **antiplatelet** agent *(Grade
    1 Level A)*

-   Consider **Clopidogrel** as the single agent (*Grade 2 Level B*) --
    CAPRIE[@capriesteeringcommittee1996]

-   **High-intensity statin therapy** to reduce all-cause and
    cardiovascular mortality - Atorvastatin 80mg or Rosuvastatin 40mg.
    These can be titrated down to atorvastatin 40mg or rosuvastatin 20mg
    if unable to tolerate. (*Grade 1 Level A*) [@arya2018; @grundy2019]

-   Control **Hypertension** to BP target \<140mm Hg systolic and \<90mm
    Hg diastolic in patients with CLTI (*Grade 1 Level B*)

-   Offer **Smoking Cessation** interventions and ask all smokers or
    former smokers about status of tobacco use every visit (*Grade 1
    Level A*)

-   Diabetic foot wounds with signs of infection, erythema, swelling,
    pain and foul smelling drainage should be investigated with plain
    radiography, ESR, CRP, cultures and, managed with iv antibiotics.
    Signs of systemic sepsis, such as fevers, tachycardia or shock, such
    as hypotension, should warrant urgent debridement and drainage,
    regardless of vascular status.

**What imaging assessment is required?**

The CLTI Guidelines outline an algorithm of attaining Arterial
Anatomical Imaging. Starting with US and then depending on the
information required, CTA, MRA, or eventually digital subtraction
angiography. They emphasize the importance of obtaining good quality
imaging to appropriately stage and be able to compare the level and
degree of disease.

For wounds with concern for underlying osteomyelitis - initial workup is
with plain radiography, which can identify soft tissue emphysema,
evidence of osteomyelitis, or presence of a foreign body. High suspicion
of early osteomyelitis with negative x-ray may warrant an
MRI.[@giurato2017]

**What is the Global Limb Anatomic Staging System (GLASS)?**
[@conteGlobalVascularGuidelines2019a]

Because the existing arterial anatomical staging of disease is vague,
"lesion focused", and not all encompassing (beyond the concept of
'in-line pulsatile flow to the foot'), **GLASS** focuses on
infrainguinal disease, and attempts to incorporate all aspects in its
staging to improve vascular care and evidence-based revascularisation
(EBR) outcomes. 

GLASS is a grading system based on anatomical and subjective assessment
of calcification and incorporates two novel and important concepts:

1.  The Target Arterial Path (TAP)

2.  Estimated Limb-Based Patency (LBP)

As GLASS focuses on Infrainguinal disease, with the aortoIliac (AI)
segment considered the inflow disease which includes the Common Femoral
Artery and the Profunda Artery. Therefore, the GLASS grades assume the
inflow vessels are treated and adequately 'dealt with'. 

Infrainguinal disease assessment for Femoropoliteal (FP) and
Infrapopliteal (IP) is based on length of disease and the extent of
CTOs. The FP and IP GLASS Grades are then combined into Stages 1-3. 

The calcification scale is a dichotomous subjective assessment of the
degree of calcification and if there is \>50% circumference of
calcification, diffuse or bulky calcification or "coral reef" plaques,
then there is an increase in the within-segment grade by one numerical
value.

There is also mention of the Inframaleolar (IM) degree of disease (PO,
P1-absent arch, P2-no target artery crossing into foot) which is not
included in the GLASS staging given little evidence on the outcomes this
difference makes on overall patency and limb salvage.

Once the GRADES (0-4) of FP and IP disease are determined then staging
(1-3) can be performed based on the matrix or grid that is provided.
Staging then allows for estimated Peripheral endoVascular Intervention
outcomes (PVI) to be predicted, Immediate Technical Failure (ITF - \<10%
or \< 20% or \> 20%)) and 1-year Limb Based Patency (LBP - \>70%, 50-75%
or \<50%). 

**What is the Target Arterial Path (TAP)?**

"The selected continuous route of in-line flow from groin to ankle. The
TAP typically involves the least diseased IP artery but may be angiosome
based."

**Which revascularisation management strategies exist for CLTI?**

The mainstay of management for patients with CLTI have always been based
on the fundamental principle of limb salvage. Given the high risk of
limb loss in these patients there is a low threshold to revascularize
these patients if they have occlusive disease that is treatable. But
strategy has varied significantly.

The CLTI Guidelines provide an approach to dealing with this complex
condition on planning three aspects to each case:

1.  **Patient Risk Estimation**

2.  **Limb Staging**

3.  **Anatomic Pattern of Disease**

**What is involved with the Patient Risk Estimation?**

***Good Practice Statements (Recommendations section 6)***

"Refer all patients with suspected CLTI to a vascular specialist for
consideration of limb salvage, unless major amputation is considered
medically urgent."

"Offer primary amputation or palliation to patients with limited life
expectancy, poor functional status (e.g. non ambulatory), or an
unsalvageable limb after shared decision-making."

***Recommendation 6.3:***

-   Estimate periprocedural risk and life expectancy in patients with
    CLTI who are candidates for revascularization. *Grade 1 (Strong)
    Level of Evidence C (Low)*

-   Average Surgical Risk: \\\<5% operative mortality and 2-year
    survival \>50%

-   Severe Surgical Risk: \\\>/= 5% operative mortality and 2-year
    survival \</=50%

Understanding disparities is important when interpreting risk for
amputation. Observational studies have found certain populations to be
at higher risk of amputation regardless of disease severity. These
groups include African Americans, lowest median income, medicaid
insurance, uninsured, or those from regions with less access to vascular
surgeons.[@hughes2019; @ho2005; @eslami2007]

**What is involved and recommended with the Limb Staging and
recommendation for Management?**

-   Use an integrated threatened limb classification system (such as
    WIfI) to stage all CLTI patients who are candidates for limb
    salvage. *Grade 1 (Strong) Level of Evidence C (Low)*

-   Perform urgent surgical drainage and debridement (including minor
    amputation if needed) and commence antibiotic treatment in all
    patients with suspected CLTI who present with deep space foot
    infection or wet gangrene. (Good Practice Statement)

-   Offer Revascularisation to all "average surgical risk patients"
    (\\\<5% operative mortality and 2-year survival \>50%) with advanced
    limb-threatening conditions (e.g. WIfI stage 4) and significant
    perfusion deficits (e.g. ischemia grades 2 and 3). Particularly if
    they have good saphenous vein. *Grade 1 (Strong) Level of Evidence C
    (Low)* [@adam2005; @norgren2007]

**What is involved in the Planning of the Anatomic pattern of disease
and its effects of revascularisation strategy?**

The overall pattern of arterial occlusive disease is a dominant factor
in guiding type of revascularisation and timing of such.

**Do all patients require direct in-line flow to the foot as a primary
technical outcome with revascularisation?**

Patients with rest pain do not necessarily require direct in line flow
are those with rest pain "for which correction of inflow disease alone
or treatment of FP disease even without continuous tibial runoff to the
foot may provide relief of symptoms. This may also be the case in
patients presenting with minor degrees of tissue loss."

**What are some essential Key Factors to consider before deciding Open
versus Endovascular according to the CLTI guidelines?**

1.  **The "availability of and quality of autogenous vein conduit"7**

2.  **Patient overall risk (as mentioned above) and Limb Staging**

3.  **The Target EndoVascular Intervention (TVI) outcomes**

**What evidence do we have for deciding between Endo and Open
management? What is the BASIL trial?**

The evidence is largely of poor quality, and is largely retrospective or
non-controlled, or industry sponsored. **BASIL** (Bypass versus
Angioplasty in Severe Ischemia of the Leg) remains the only multicenter
RCT (BASIL -2 and 3 underway) directly comparing an endo versus open
management strategy for CLTI and Infra-inguinal occlusive disease. 

BASIL compared POBA and Bypass across multiple centers (27 centers,
n=452, 1999-2004) in the UK. Primary endpoint was Amputation-Free
Survival (**AFS**). [@bradburyBypassAngioplastySevere2005]

Major Findings:

1.  At 6-months follow up: no difference in AFS.

2.  Intention-To-Treat Analysis of overall follow up showed no
    significant difference in AFS and overall survival.

3.  Among patients who survived \>2 years, overall survival was better
    for those treated with Bypass as a first approach

4.  Analysis to treat:

    -   Prosthetic Bypass Patients did very poorly (even compared to
        POBA).

    -   Patients who had bypass after failed POBA had significantly
        worse AFS compared to those treated with a bypass as initial
        treatment.

Criticism of BASIL:

1.  Majority had POBA alone (not currently best endovascular option)

2.  25 % of Open Bypass were Prosthetic

3.  The technology and technical Skill with growing operator experience
    in Endovascular has improved.

For isolated tibial disease, first line endovascular treatment of choice
is transluminal, or subintimal angioplasty. Atherectomy, stenting, and
drug coated balloons are often used, but should currently not be
considered first line, but this data is rapidly evolving. Patency is
often poor but limb salvage is reasonable.[@popplewell2019; @kayssi2016;
@mustapha2016]

Optimal follow up for endovascular interventions of the lower extremity
has not yet been established, but should at least include a pulse exam,
ABPI and duplex to establish a new baseline after
intervention.[@zierler2018; @mohler2012]

**What are the BASIL 2, BASIL 3, and BEST-CLI Trials?**

**BASIL 2:** Infrapopliteal Disease: Vein Bypass First vs. Best
Endovascular Treatment first.

**BASIL 3:** PBA +/- BMS vs. DCB +/- BMS vs. DES

(Both Follow up 24-60 months, Primary Endpoint AFS)

**BEST-CLI:** Open Bypass versus Endovascular Intervention, Primary
Endpoint: MALE-Free Survival. Major Above-the-Ankle Amputation, Major
Bypass or Jump/interposition graft revision or the need for thrombectomy
or thrombolysis (MALE).

**Some important take away lessons for unique scenarios in operative
planning for CLTI based on anatomy and conduit.**

-   Single segment GSV best conduit for infrageniculate bypass.
    [@arvela2010; @avgerinos2015; @moreira2016]

-   Fem-AK pop bypass with prosthetic may be preferred to contralateral
    GSV. [@moreira2016] If the target is below the knee, then all
    autogenous conduits--contralateral GSV, SSV, and spliced arm
    vein--are preferred over prosthetic grafts.[@brochadoneto2014;
    @faries2000; @taylor1987]

-   Bilateral external iliac occlusion may be best treated with end to
    side aorto-bifemoral bypass to allow for continued perfusion of the
    pelvis. [@jaquinandi2008; @vandenakker1992; @brewster1978]

-   Profunda-popliteal collateral index (Segmental pressures AK-BK/AK)
    of less than 0.25 may suggest that there is sufficient collateral
    network between profunda and popliteal that SFA treatment may not be
    necessary. [@boren1980; @mawatari2000]

-   Planning distal bypasses should take into account the angiosome of
    the wound and the most distal healthy inflow vessel to result in the
    shortest bypass possible.[@hingorani2016; @jongsma2017]

-   Intensified anti-thrombotic therapy may be needed in patients with
    "high risk" infrainguinal bypasses--prosthetic conduit, below the
    knee targget, suboptimal conduit, poor arterial runoff, extensive
    lesions or tissue loss. Single anti-platelet may not be sufficient,
    and should be intensified to asa/rivaroxaban, dual antiplatelet, or
    anticoagulation with Vit K antagonist.[@conte2015; @hussain2018;
    @strobl2013; @tepe2012]

#### Complications

**What are some complications of lower extremity revascularization
procedures?**

For a comprehensive list of access complications after endovascular
therapy, see \@ref(access-comp).

Femoral exploration carries high risk of infection or lymphatic leak.
Lymphatic leaks often resolve spontaneously. Infected lymphatic leaks,
particularly in the setting of prosthetic bypass require exploration.
Should a lymph leak not resolve with conservative management, definitive
therapy requires alcohol ablation or muscle flap coverage. The most
important aspect of lymph leak is prevention with careful dissection and
tissue management during femoral exploration.[@obara2014; @weaver2014]

## Acute Limb Ischemia

**Accompanying Episode:** Acute Limb Ischemia: Alex Forsyth and Sarah
Carlson

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/VSITE_ALI_postauphonic.mp3")

```

**What is Acute Limb Ischemia and what does it encompass?**

Acute Limb Ischemia (ALI) is defined as any process that leads to an
abrupt cessation of blood flow to a limb resulting in ischemia. There
are several causes, but the most common two are embolic and thrombotic.

-   **Embolism**

    -   ***Cardiac***

        -   Typically due to atrial fibrillation.

        -   Acute upper limb ischemia is most commonly due to cardiac
            embolism. 

        -   Endocarditis -- as seen in IV drug users or patients with
            bacteremia from other causes.

        -   Cardiac tumors -- such as atrial myxoma.

    -   ***Atherosclerosis*** -- e.g. iliac disease embolizing
        downstream to the lower leg.

    -   ***Paradoxical Embolism*** -- Thromboembolic venous system with
        PFO. 

    -   ***Aneurysm*** -- e.g. thrombus from within an aortic aneurysm
        embolizing downstream to the leg.

-   **Thrombosis**

    -   ***Aneurysm***

    -   ***Bypass Graft***

    -   ***Acute on Chronic Progression of Atherosclerosis***

    -   ***Dissection***

    -   ***Thoracic Outlet Syndrome*** -- in the upper extremity

    -   ***Vasospasm*** (severe)

For clarification:

1.  Acute on Chronic Progression of Atherosclerosis: Once a chronic
    stenosis becomes critically tight, platelet thrombus can develop
    leading to an acute occlusion; or unstable plaque can "rupture"
    leading to an acute occlusion of a chronic lesion.

2.  Regarding Aneurysms: Small aneurysms (such as popliteal) are less
    likely to rupture, but more likely to thrombose and cause an acute
    limb ischemic event.

### Presentation and Diagnosis

**What is the patient presentation of ALI? Are there any differences
between upper and lower extremity presentations?**

Classically remembered by the 5 or 6 Ps (depending on who you ask)

-   ***Pain***: Usually located distal to the occlusion and gradually
    increases in severity as the ischemic time increases until
    neuroischemic sensory loss occurs.

-   ***Pallor***: the limb appears pale compared to the non ischemic
    limb. There is delayed or absent capillary refill.

-   ***Poikiolothermia***: (just a way to make "cold" into a "P" --
    really means cold limb) Literally the inability to regulate one's
    body temperature, or dependent on ambient temperature as cold
    blooded animals are. If there is no limb perfusion, it acclimates to
    the ambient temperature and feels cold to the touch. 

-   ***Pulseless***: Self explanatory. It is worth examining the
    contralateral limb, if it has normal pulses, it suggests the absence
    of PAOD or CLTI making thrombosis in situ less likely to be the
    cause of ALI. 

-   ***Paresthesia and Paralysis*** are the last two Ps. Paresthesia is
    an earlier sign of ischemic nerve dysfunction, and paralysis is a
    later sign. In the lower leg, ischemic changes often affects the
    anterior compartment first, and sensory loss over the dorsum of the
    foot is, therefore, one of the earlier neurologic deficits in ALI.

This is why a thorough physical exam is key; comparing both limbs and
including handheld doppler exam. It can be difficult to tell whether a
limb is acutely threatened, especially in patients with chronic disease
where the presentation of an acute change can be more subtle. This is
why it is also essential to assess sensory and motor function in
addition to the pulse exam. This helps gauge the urgency of
intervention.

Acute paralysis, mottling of bilateral lower extremities and absent
femoral pulses should raise concern for aortic occlusion.[@wang2016]

**How is ALI classified?**

From Rutherford RB, Baker JD, Ernst C, et  al. Recommended standards for
reports dealing with lower extremity ischemia: revised version. J Vasc
Surg. 1997;26:517--538. [@rutherfordRecommendedStandardsReports1997]

1.  Viable - Not immediately threatened

    -   Sensory Loss: None; Motor Weakness: None

    -   Arterial Signal: Present; Venous Signal; Present

2.  Threatened

    a\. Marginally - Salvageable if promptly treated (\<24hrs)

    -   Sensory Loss: Minimal (toes) or none; Motor Weakness: None

    -   Arterial Signal: Weak/absent; Venous Signal: Present

    b\. Immediately - Salvageable if immediately treated (\<6hrs)

    -   Sensory Loss: More than toes; Motor Weakness: Mild/Moderate

    -   Arterial Signal: Absent; Venous Signal: Absent

3.  Irreversible - Major tissue loss or permanent nerve damage
    inevitable; Minimal benefit from revasc

    -   Sensory Loss: Profound anesthesia; Motor Weakness: Dense
        paralysis (rigor)

    -   Arterial Signal: Absent; Venous Signal: Absent

**What does the work up for ALI entail? How is the diagnosis made?**

-   The diagnosis can often be made on history, physical exam, and
    bilateral ABPIs. Imaging can be done in patients in who the
    diagnosis is uncertain.

-   As with most urgent cases, the type of imaging done depends on the
    availability at your institution, but generally CT angiography or
    arteriography should be done on viable and marginally threatened
    limbs. Arteriography often can distinguish between embolic vs
    arterial thrombosis which may help to direct management.

-   Bedside ultrasound can also be very helpful, especially color flow
    doppler US.

The decision to perform cross-sectional imaging before angiography
varies depending on how severe the presentation is and how quickly you
can obtain imaging. The patient's renal function may also influence the
decision to administer two contrast loads (CT followed by endovascular
intervention). If a femoral pulse is present it could be feasible to
proceed with on-table angiogram without a CT scan. If femoral pulses are
absent there is increased suspicion of aortoiliac disease and therefore
CT angiogram will be beneficial for both diagnosis and operative
planning.

In patients with severe renal insufficiency, MRA or MR time-of-flight
can be helpful, but these studies usually take longer to obtain and may
not be quickly available in an acute threatened limb situation.
Therefore going straight to angiography should be considered, as to get
the most immediate and best imaging of the tibial vessels.
[@creagerClinicalPracticeAcute2012; @earnshaw100AcuteLimb2019]

### Management

**What is normally done in the initial management of ALI?**

-   ***Anticoagulation***: IV unfractionated heparin should be
    immediately administered to prevent proximal and distal progression
    of secondary thrombus as long as heparin is not contraindicated. The
    dose should be titrated to maintain activated partial thromboplastin
    time between 50 and 80 seconds (2-3 times normal values). A good
    rule of thumb for IV heparin is to start with a bolus of 80-100
    units/kg, and then drop at 18units/kg/hr -- titrating to PTT at 2-3x
    normal -- or follow local protocol.

-   ***Supportive Care***: Analgesia, iv fluids.

-   ***Labs***: A full panel including serum chemistry with BUN, Cr,
    CBC, and coagulation studies should be obtained. Baseline plasma CPK
    can be helpful to monitor for evidence of rhabdomyolysis after
    reperfusion.

**What are some of the options for treatment of ALI?**

-   **Medical**: Primarily with anticoagulation using heparin or a
    direct Xa inhibitor.

-   **Open Surgical Management**:

    -   ***Thrombectomy***: Balloon catheter based (Fogarty embolectomy
        balloon -- Dr. Fogarty invented this while he was a medical
        student)

        -   Patients with acute limb ischemia, neuro deficit and distal
            pop embolism may be best treated with popliteal exposure and
            open thrombectomy. [@darwoodSurgeryThrombolysisInitial2018;
            @kempeResultsSurgicalManagement2014]

    -   ***Bypass***

        -   Patients with aortic occlusion and significant
            comorbidities, and significant aortoiliac baseline disease
            may likely benefit from primary axillo-bifemoral bypass.
            [@mohapatraRiskFactorsPerioperative2018]

    -   ***Endarterectomy*** -- not necessarily required but might be
        used for common femoral exposure and distal access depending on
        vessel disease burden.

-   **Endovascular**

    -   ***Pharmacological Catheter Directed Thrombolysis***

        -   Absolute contraindications include active bleeding disorder,
            CVA (\<6mo), CNS injury or head injury (\<3mo), or GI bleed
            (\<10d).

        -   Relative contraindications include recent major surgery,
            uncontrolled hypertension, intracranial tumor, pregnancy,
            recent eye surgery, hepatic failure, CPR (\<10d), or
            bacterial endocarditis.

        -   1-2% risk of hemorrhagic stroke. If neurologic deficit
            develops during thrombolysis, stop lysis and perform
            immediate CT head.
            [@ourielComparisonRecombinantUrokinase1998]

        -   If fibrinogen drops below 100mg/dl, recommend immediate
            cessation of lytic agent and consider cyroprecipitate
            (includes fibrinogen, factor VIII, von Willebrand's factor,
            factor XIII).[@mann2019]

    -   ***Percutaneous Thrombus Aspiration***: Useful for small fresh
        thrombi such as after angioplasty, as distal diameter of the
        catheter tip limits the size of the thrombus that can be
        removed. 

    -   ***Mechanical Thrombolysis and Aspiration***: Are also useful
        for patients with contraindications for thrombolytic therapy,
        and also may allow for a lower dose of a thrombolytic agent, but
        risk damage to the arterial wall.

**Who gets which kinds of treatment? Who needs emergent treatment?**

-   **Class I** patients might just need medical therapy like
    anticoagulation and revascularization can be elective. 

-   **Class IIb** patients do not need immediate revascularization

    -   If symptoms have been present for less than 2 weeks endovascular
        therapy is preferred

    -   If more than 2 weeks or lytic therapy has failed then surgical
        intervention is preferred 

-   **Class IIb** need immediate revascularization. Historically
    surgical revascularization has been preferred because of its
    immediacy, but catheter directed thrombolysis and percutaneous
    mechanical thrombectomy have shortened time to revascularization. 

-   Related Studies:

    -   ***STILE Trial***: One of the first large RCTs comparing
        catheter thrombolysis with open surgery. Overall the study
        showed some short term benefit to open surgery however this can
        probably be attributed to: (1) in 28% of patients randomized to
        CDT they weren't able to get a catheter in place so these
        patients were considered treatment failures and crossed over to
        the surgery arm, and (2) patients with very long duration of
        ischemia, up to 6 months, were also included in the study. When
        outcomes for patients who had been symptomatic for less than 2
        weeks were analysed, the thrombolysis patients actually did
        better. [@thestileinvestigatorsResultsProspectiveRandomized1994]

    -   ***TOPAS Trial***: Larger RCT which enrolled patients who had an
        acute arterial occlusion of less than 14 days. This showed no
        difference in mortality or amputation-free survival between the
        open surgery and CDT groups but higher major bleeding in the CDT
        group. [@ourielComparisonRecombinantUrokinase1998]

    -   ***Meta Analysis***: Originally published in 2002 but updated in
        2013 and 2018. Demonstrated no difference in mortality or limb
        salvage between surgical and thrombolytic therapy, but
        endovascular demonstrated higher rates of complications
        including ongoing limb ischemia and bleeding within 30 days of
        treatment. Previously they had reported higher rates of stroke
        in the thrombolysis category but the most recent update is
        unable to support this finding.
        [@darwoodSurgeryThrombolysisInitial2018]

        -   Evidence considered it is very reasonable to think about a
            catheter-directed therapy especially if the presentation is
            acute, less than 2 weeks or so. There are, however, certain
            anatomical locations that most surgeons would favor a simple
            open procedure such as balloon thrombectomy , such as
            embolism to the common femoral or brachial arteries.   

-   **Class III** ALI is usually treated with primary amputation because
    revascularization is unlikely to restore function to the limb and
    restoring blood flow can cause the patient serious harm. 

**What are the risks of revascularization for a class III or prolonged
ischemia?**

Myonephropathic metabolic syndrome: muscle cells undergo liquefaction
necrosis due to ischemia. Potassium, myoglobin, lactic acid, and
superoxide accumulate and can perfuse through the body or can have a
sudden increase in the event of revascularization which leads to
hyperkalemia, arrhythmias, pulmonary edema, metabolic acidosis,
myoglobinuria, and can even cause sudden death from heart and/or renal
failure. [@obaraAcuteLimbIschemia2018] Treatment of this is largely
supportive with fluids.

## Compartment Syndrome

**What is the pathophysiology and presentation of compartment
syndrome?**

-   Increased intramuscular compartment pressure results from increases
    in capillary permeability due to ischemic reperfusion. The increase
    in pressure leads to neuromuscular dysfunction and interferes with
    circulation. Irreversible damage occurs when pressures exceed 30mmHg
    in each compartment.  

-   Ischemia reperfusion causes increased capillary permeability due to
    free oxygen radicals, neutrophils and endothelial factors that
    collect during ischemia. Severity depends on time to reperfusion,
    muscle mass, and flow pattern (i.e. direct or collateral). Symptoms
    often develop within 6 hours of reperfusion. [@vonkeudell2015]

-   The pathophysiology underlying acute extremity compartment syndrome
    is related to the arteriovenous pressure gradient theory. Increased
    compartment pressure reduces the gradient, reducing arterial
    pressure and increasing venous pressure.[@frink2010; @mcqueen1996;
    @elliott2003; @papalambros1989]

-   Clinically patients can have neurological dysfunction with sensory
    motor deficits, but the most common presentation is a tense
    extremity with severe pain on passive movement of the muscles in the
    compartment. If the lower leg is affected this is often on
    dorsiflexion/plantar flexion of the ankle. A sensitive indicator is
    loss of two point discrimination (proprioception).

**How do you diagnose compartment syndrome?**

-   **Physical Exam**: Tenderness, (especially over anterior
    compartment) and paresthesiae, (especially between first and second
    toes).

    -   ***Anterior Compartment***: Sensory distribution and muscular
        innervation from the deep peroneal nerve (this is a VSITE
        favorite).

    -   ***Deep Posterior Compartment***: Sensory distribution and
        muscular innervation from the tibial nerve[@velmahos2002].

-   **Compartment Pressures**:

    -   A needle is required to access the compartment and a pressure
        monitoring system (this can be handheld Stryker kit, or just a
        hollow bore needle connected to an arterial pressure bag).

    -   ***Normal compartment pressure is \<10-20mmHg***; greater than
        30 is highly concerning.  More accurate than an absolute number
        is comparing the compartment pressure to the mean arterial
        pressure or diastolic pressure.  If the compartment pressure is
        within 40mmHg of the MAP (for example, MAP is 60 and compartment
        pressure is 25 -- this is concerning) -- OR -- if the difference
        between compartment pressure and diastolic pressure is less than
        10 (for example, diastolic pressure is low at 30, and
        compartment pressure is 22).

If the clinical suspicion of compartment syndrome is high, do not be
reassured by "normal" compartment pressures.  It is relatively low risk
to do fasciotomies, but the risk of limb loss is high for a missed
compartment syndrome.

**Management of compartment syndrome is a fasciotomy, what are the types
of fasciotomies?**

-   **Forearm and Upper Arm Fasciotomies**: Often performed by
    orthopedic or hand surgeons. The forearm fasciotomy includes dorsal
    and volar incisions to release the dorsal and volar compartment, and
    mobile wad, while avoiding numerous superficial cutaneous nerves.
    The arm fasciotomy releases the medial, lateral, and deltoid
    compartments through medial lateral incisions.

-   **Thigh**: Performed through medial and lateral incisions to release
    the lateral, medial and posterior compartment.

-   **Lower Leg**: The most common type is a lower leg 4 compartment
    fasciotomy and is explained in more detail below.  

**How is a lower leg 4 compartment fasciotomy for the lower extremity
performed?**

1.  A longitudinal incision is created between the fibular shaft and the
    crest of the tibia over the intermuscular septum and the anterior
    and lateral compartments are opened. If tissues are swollen
    occluding the view of the intermuscular septum, the perforating
    vessels can be followed down to it. Nerves including the peroneal
    nerve are most at risk near the fibular head.

2.  A second incision is created on the medial surface of the lower leg
    approx. 1cm posterior to the edge of the tibia to avoid the greater
    saphenous vein. The superficial posterior compartment is incised.
    The gastrocnemius-soleus complex is taken down from its attachments
    to the tibia in order to access the deep posterior compartment. 

3.  The incisions should be made generously. Sometimes the skin incision
    can be a little short of the fascial incision, but they should be
    long in order to fully release the compartments.

4.  After hemostasis loose dressings should be applied. The leg should
    be elevated to reduce edema that can complicate closure. Closure can
    be done in 48-72 hours, but may be delayed and dressed with wound
    vacs to attempt primary closure. If primary closure is not possible,
    a split thickness skin graft can be used.

**Who should a prophylactic fasciotomy be performed on?**

-   Patients with high occlusion and extensive ischemia.  

-   Acute ischemia of greater than 6 hours with few collaterals.

-   Patients with combined arterial and venous injury.

-   Patients who are obtunded making serial examination difficult.

**What is the prognosis for a patient with ALI? What are some patient
factors that lead to a poor prognosis?**

Amputation rates after acute limb ischemia are typically described in
the 10-20% range, and mortality is also in the 10-25% range whether
you're talking about surgery or catheter-directed procedures (that's
excluding the patients who present with Rutherford class III and by
definition have an unsalvageable limb). Many factors determine
likelihood of amputation; typically, patients with more medical comorbid
conditions tend to do worse as you might expect: baseline CAD, kidney
disease and smoking are predictive of worse outcomes. There is a trend
toward improved limb salvage rates (decreased amputation rate) over
time, and I think this speaks to wider availability of different limb
salvage techniques among vascular surgeons across the globe.

## Blue Toe Syndrome

### Etiology and Presentation

Blue toe syndrome is characterized as a painful discolored toe in the
setting of a normal vascular examination or palpable pulses.

Acute onset of blue toe after MI - consider cholesterol embolization
from wire manipulation in the aorta.
[@saricAorticAtherosclerosisEmbolic2012]

### Evaluation

Patients should have a work up for investigation of an embolic source.
Patients without a clear recent source of embolization on history or
physical examination, should undergo:

-   CTA of chest/abdomen/pelvis to look for an arterial lesion.

-   Echocardiogram to identify a cardiac source.

-   Holter monitor to evaluate for an underlying arrhythmias.

If no obvious source of embolism can be identified, further
investigation for underlying cancer may be required due to malignancy
leading to prothombotic state.

### Management

**Cholesterol Embolization**: Medical management with a single
antiplatelet agent and statin. Anticoagulation may precipitate further
embolization. [@ghahramaniPostproceduralBlueToes2016;
@quinonesCholesterolEmboliSyndrome2013]

**Specific Thrombogenic Arterial Lesion**: Medical management with
anticoagulation. Recurrent symptoms on anticoagulation may require stent
coverage or, in a young patient, thrombectomy.
[@reyesvaldiviaAnticoagulationAloneAortic2017;
@jeyabalanEndovascularStrategiesTreatment2014;
@vermaContemporaryManagementSymptomatic2014]

## Non-atheromatous Popliteal Artery Disease

### Popliteal Artery Entrapment Syndrome (PAES)

#### Etiology and Presentation

Most often seen in young healthy patients (often athletes) who present
with claudication.

Etiologically due to embryologically abnormal lateral attachment of the
medial head of the gastrocnemius muscle.[@gokkus2014; @lejay2014] Type
IV PAES caused by compression with popliteus muscle.

#### Evaluation

Physical examination findings will be loss of pedal pulse on active
plantar flexion or passive dorsiflexion of the foot.[@gokkus2014;
@lejay2014]

Best diagnosed with axial cross-sectional imaging from the MRA to
identify the abnormal muscle course and insertion. [@sinha2012;
@lejay2014]

#### Management

Symptomatic PAES should be treated, even if presenting with mild or
moderate claudication, due to risk of progression and thrombosis from
scaring.[@forbes2019]

### Cystic Adventitial Disease

#### Etiology and Presentation

Controversial etiology, often considered related to repeated popliteal
trauma due to knee flexion, causing fluid to collect between the
adventia and media layers of the artery wall. Presentation overlaps with
atherosclerotic lesions of the popliteal segment, but in patients
without traditional risk factors.

#### Evaluation

Often diagnosed on duplex ultrasound or CTA. For a representative image,
see \@ref(cad).

#### Management

Compression can sometimes be relieved by percutaneous drainage of large
cystic portion that is causing luminal compression. PTA and stenting
should be avoided as often poor durability. Operative resection and
reconstruction with an interposition graft can be effective but should
be considered as a final management strategy.[@li2017]

## Amputations

The majority of questions regarding amputations stem from decision
making around peripheral arterial disease. You may not see many
questions asking uniquely about amputation technique or management,
however this topic is VERY important for the successful management of
peripheral arterial disease. Rutherford's has two fantastic chapters on
decision making[@bianchi2018] and techniques[@eidt2019].

The most commonly performed lower extremity amputations include:

### Above-knee, trans-femoral amputation

Above the knee amputations are most often performed with a fish mouth
incision. Suture ligation of the neurovascular bundle at the level of
the mid SFA and transection of the femur.

### Below-knee, trans-tibial amputation

Below the knee amputation often performed with a long posterior, or
Burgess, flap. Each tibial neuro vascular bundle is ligated
individually. The fibula is transected 2-3cm more proximally than the
tibia to avoid weight bearing from the prosthetic.

Although hotly debated, a Cochrane review demonstrated no difference in
outcomes based on flap type. Although it does recommend performing in
two stages in the setting of wet gangrene.[@tisi2014]

### Transmetatarsal amputation

Transmetatersal amputation is performed with transection of the
metatarsal bones just distal to the heads. Best to transect in a
graduated parabola with the 5th metatarsal most proximal. Patients may
benefit from concomitant Achilles tendon lengthening to prevent equinous
deformity that can result in ulceration of the medial aspect of the
TMA.[@eidt2019]
