---
editor_options: 
  markdown: 
    wrap: 72
---

# Vascular Trauma

## Peripheral

**22 Sep 2019**: *Kevin Kniery, MD, MPH; Todd Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Rasmussen_Final_Final.mp3")

```

**Hard and Soft Signs of Vascular Injury**

Hard signs of a peripheral vascular injury:

-   Obvious bleeding (arterial or venous)

    -   can be pulsatile or even history of bleeding in the field

-   Expanding hematoma

    -   underneath a closed wound/injury

-   Profound ischemia of the extremity

    -   determined by the absence of a palpable pulse or dopplerable
        signal beyond the area of injury

-   Presence of an audible bruit or palpable thrill near site of injury

    -   concern for an arterial-venous fistula, often high flow

Soft signs of a peripheral vascular injury:

-   Injury pattern often associated with a vascular injury (i.e.
    posterior knee dislocation)

-   Diminished blood flow to extremity

    -   weak, but audible arterial doppler signal

-   Penetrating wound in proximity to a major axial vessel

-   Peripheral nerve injuries

    -   Femoral nerve injury
        -   parasthesia to anterior thigh
        -   decreased hip flexion and knee extension
    -   Sciatic nerve injury
        -   parasthesia to lateral leg and dorsal/lateral/plantar foot
        -   weakness of foot plantar flexion
    -   Tibial nerve injury
        -   parasthesia over heel
        -   weakness of plantar flexion
    -   Deep peroneal nerve injury
        -   parasthesia of first digital interspace
        -   foot drop[@bulger2014]

**Initial principles of managing a bleeding extremity**

-   DO NOT GET DISTRACTED
    -   These patients are often poly-trauma patients with likely other
        injuries
    -   Follow ABCDEs of the primary survey
        -   Identify bleed and hold pressure to stop the bleeding
        -   Pressure can be manual pressure, such as placing a finger in
            the wound, or applying a tourniquet
        -   Once bleeding is effectively controlled, RESUSCITATE and
            continue the primary survey to address other life
            threatening injuries

**In Rich's Vascular Trauma Book, they described the injury extremity
index. Could you explain this concept?**

If I had to take one tool with me into resuscitation room or an austere
location to evaluate for vascular injury, it would be the continuous
wave Doppler. It's only more powerful if you can combine it with a
manual blood pressure cuff and together allow one to obtain an objective
measure of perfusion. One can slowly inflate the blood pressure cuff,
just proximal to the arterial signal, kind of like taking a blood
pressure, and measure the pressure at which the arterial signal goes
away. That can then be compared to a normal or non-injured extremity. If
there is no flow limiting arterial injury, the ratio should be one or
slightly greater than one.

Conversely, if the ratio for the injured extremity is less than 0.9,
then that's indicative of flow limiting injury in the artery to that
extremity. If the patient happens to be in shock or cold and clamped
down, the injured extremity index can be repeated 10 or 15 minutes later
as the patient is resuscitated and warmed.[@rasmussen2022]

**So what do you think the role of CTA is in the workup of extremity
with concern for vascular trauma?**

If there are hard signs of arterial injury, oftentimes the patient will
require an intervention without obtaining formal imaging. The question
about imaging often comes into play when there's a soft sign of vascular
injury, such as a reduced injury extremity index, audible bruit, or
injury pattern. We spent decades saying that the traditional
arteriography was the gold standard, but I think we can say now that CTA
has become the gold standard for the evaluation of extremity vascular
injury. CTA has improved in quality and it can also get imaging of the
head, torso, or other other extremities.

**And so whether the patient has a hard sign or a soft sign with a
vascular injury, can you talk us through some of the principles of
operating on peripheral vascular trauma?**

Foremost important is to resuscitate the patient and to be mindful of
his or her physiology and hemodynamics. As we said, an extremity
vascular injury can be somewhat distracting. Make sure the patient is
being resuscitated and warmed, ventilator settings are being corrected,
and monitoring is in place -- which takes a lot of communication with
the anesthesia team, the circulating nurse, the technicians, and the
blood bank.

I think once the patient is set and you have the team resuscitating him
or her, then you can turn your attention to the injured extremity. I
will prep that patient widely and be mindful that the incision is likely
to take more than what you think it will. For example, for an above the
knee popliteal artery, I will still prep that patient from their
umbilicus down to both legs; because you need to anticipate getting
proximal control, or doing an angiogram at the femoral artery level and
anticipate the potential harvesting of saphenous vein for conduit.

I also leave the tourniquet on and just prep it into the case. Taking
that tourniquet down prematurely can be problematic for a couple of
reasons. It may result in arterial bleeding, which in itself can be
catestrophic, but it can also result in hypotension which poses risk to
multiple organ systems. So, I just prep them into the field and then
communicate with anesthesia to tell them when I am going to let this
tourniquet down. Prior to doing so, I ensure they are ready, they have
blood going, and have a secondary measure to control bleeding as you let
that tourniquet down.

For the incisions, they just need to be probably twice as long as you
think they do. You can't fix what you can't see and if you can't see it,
you can't control it, and you certainly can't repair it.

**Do you routinely heparinize all these patients?**

I hate to equivocate, but it depends. It depends on whether or not they
have other injuries. So if it's an isolated extremity injury and the
patient does not have any torso bleeding, head trauma, or very extensive
soft tissue injury, then I will then be more inclined to use systemic
heparin during revascularization. If they have any of the above
injuries, then I will use regional heparin flushed up and down the
injured vessel.[@liang2016; @fox2012]

**One question that I've heard come up a handful of times is can you use
saphenous vein from the injured extremity?**

Our teaching, as you referenced, has always been to use a saphenous vein
from the contralateral extremity, but I am quite practical about it. I
certainly have used ipsilateral saphenous vein. If there's no deep vein
injury and it's part of the incision and exposure, then I will use
ipsilateral reversed greater saphenous vein. If there is concern for
deep venous injury or if I have another person to harvest and who can
prep the other leg, then often I'll defer to that and use contralateral.
I'll be more deferential to using the contralateral saphenous vein in
the reverse configuration.[@liang2016; @fox2012]

**So you have a vascular injury and you have debride it back to healthy
tissue. How do you make the decision between, primary repair, patch
repair, or interposition graft with autologous vein or prosthetic?**

You know, you can picture it now, you have vascular control with clamp
proximal and distal. Now you're setting up, you are really shifting
gears into a different phase of this operation. It's important to just
take a deep breath, I think at this point, and check with anesthesia and
make sure the patient's doing all right from a global standpoint.

I think the other thing is that at this point, I have sometimes
overlooked the importance of performing a proximal and distal
thrombectomy using the Fogarty catheter. I'll make sure that I've passed
a Fogarty catheter distally to completely clear the outflow of thrombus.
Then I'll use regional heparin, 40-60cc of heparin saline down the
outflow before I apply the clamp. I'll do the same in the inflow and
that buys you a little time as you're sort of thinking about and sizing
up what's in front of you.[@liang2016; @fox2012]

I think in my experience, the majority of patients need an interposition
graft. Certainly, if it's a grazing wound or a stab wound from a knife,
there may be a role for primary repair or patch. But if one thinks
there's going to be any compromise of the arterial lumen or tension on
the repair, then you're better off with an interposition. [@liang2016;
@fox2012]

There are some special cases where primary repair might be appropriate.
Veins are much more compliant and you can often get away with a lateral
venorrhaphy. Then sometimes the upper extremities, such as the brachial
artery, has more redundancy. If it's a very focal injury and you can get
a little stretch, then you will often be able to get the brachial artery
back together primarily.

I suspect in the majority of my experience, vascular injuries require an
interposition graft. Then in regards to conduit, because most of the
wartime injuries were contaminated and quite dirty, we deferred to
autologous saphenous vein the vast majority of the time. However, if you
are in a pinch, can't find vein or don't have time to harvest vein, the
wound isn't too contaminated, and the injury is proximal, axillary or
subclavian, proximal femoral, or iliac, then I think using a prosthetic,
either Dacron or expanded PTFE Gore-Tex is also acceptable. Some recent
civilian literature says PTFE may be better than vein, but that has not
been our experience in the deployed setting.

**Now that we have covered some basic principles, lets dive in a little
deeper on each kind of peripheral vascular injury. First we're going to
talk about junctional hemorrhage, which is an injury in the junctional
area between the torso and extremity, such as the distal external
iliac/common femoral or subclavian/axillary region. These are some our
biggest fears and most difficult to deal with. Can you take us through a
hematoma in the groin or pelvis just above inguinal ligament--what are
the best ways to get control of this?**

For the iliac artery or even just the proximal common femoral, I think a
retroperitoneal exposure to the lower quadrant of the abdomen is
preferred. We sometimes refer to that as a transplant incision, which is
quite suitable depending upon the injury pattern.

As I alluded to before, don't underestimate the exposure you're going to
need. So, if it truly is a groin hematoma, think that this could be
common femoral or external iliac and in those cases, either a transplant
incision in a retroperitoneal fashion to get that controlled up to the
common iliac, or just doing a laparotomy and finding it through the
abdomen would suffice.

**What if you encountered an internal iliac injury deep in the pelvis
that you weren't able to control, what are the consequences of ligating
that?**

That's real tiger country, in those cases it's really damaged control.
One of the things we haven't mentioned is the important tenant -- that
maybe arguably is one of the most important tenants in vascular damage
control -- is ligation. We've talked a lot about setting up for a
vascular repair, but one of the reasons to check with anesthesia and to
assess the whole of the patient is to recognize that in some scenarios
the best thing is ligation and then dealing with whatever consequences
may result from that.

Keeping in mind, the idea of life over limb and not being too aggressive
about trying to do some sort of exposure and repair of a vessel that
takes hours and many, many units of blood. So in that context, I think,
if you find bleeding from the internal iliac or hypogastric vessels, I
think ligating those is quite acceptable. There can be instances of
pelvic ischemia, but if the contralateral iliac artery is open, then
that's quite rare.

Also, often one is dealing with artery and vein injury because they're
right there together. And as I mentioned, that's real tiger country. If
that's happening in those situations, keep in mind the tenants of
exposure, light, two suctions, and don't use too small of a needle.
That's where the SH blunt tip needle on a 3-0 or 4-0 prolene needle
works well. Don't try to be too finesse about this because, in the
pelvis, you won't be able to see your needle.

**Let's take the same injury and move it now to the external iliac--what
are the consequences of ligating the external iliac artery and what are
some of your options?**

Because it's the axial vessel to that extremity, I would not ligate the
external iliac artery with such impunity. I think in those situations,
controlling it and making that assessment of whether or not to repair it
is paramount. I think that the consequences of ligating an external
iliac artery are going to be significant and include proximal extremity
ischemia, including thigh, and probable need for above the knee
amputation and even disarticulation of the hip.

So I think the external iliac needs to be preserved at all costs. And
whether you reconstruct immediately or use a temporary vascular shunt as
a damage control is dependent upon how the patient's doing or if you are
in the field and there are other triage concerns.

**Some can find shunts a little tricky, can you give some advice to
listeners on how to best use temporary vascular shunts for damage
control?**

I think it's important to know what shunts you have available before you
are in this situation. So stroll through the operating room or your
stock room, wherever you are, when it's not busy and check out the stock
of shunts. There are four or five different types of shunts, anything
from what's referred to as the Javid shunt, Sundt shunt, or there's the
Argyle shunt. Then there can be makeshift shunts, such as small caliber
chest tubes, which are temporary plastic tubes that can be placed into
the injured vessel proximally, and then controlled with either a
ligature around the vessel to lock that shunt place or a rubber vessel
loop.

And then you want to make sure that you did not blow out any clot with
that shunt. So you let some heartbeats go with red blood running through
the shunt, then occlude it with the clamp, and then insert it to the
vessel that is distal to the injury.

Likewise, you have to make sure that the thrombus has been removed, and
I mentioned this earlier from that distal vessel. I think it is very
important before you place the shunt in that you try to clear as much of
the thrombus as you can, proximal and distal, and then place that shunt
in and secure it either with a heavy silk tie or plastic vessel loops.

Most of those shunts you can listen with the Doppler and you will see
arterial flow in it. So you can listen on the surface of the plastic
shunt with a little water or acoustic gel and hear the arterial flow in
those shunts.

**For our military docs out there, if they're deploying to a role and
their supply seargent calls them and asks what kind of shunts do you
want us to have --- what sizes and what types? What recommendations
would you give to make sure that you have them on hand?**

Yeah, it's a great question for now and it's a challenge for us in the
future. I think here and now, I would try to really make sure that the
surgeon put his or her hands on Argyle shunts. The Argyle shunts come
four in a container, there's a 12 French, 14 French, 10 French, and an 8
French all sort of in the same container. And its important to make sure
he has those. As the larger 14 French Argyle can work on an iliac. So
that would be one, and another would be to look for a 14 French chest
tube. And I would say, "so these are my shunt options and I would need
to have these available."

The other shunts, the Sundt and Javid shunts, are the other ones you
could ask for, but then you only have one size for those. And so those
and the thrombectomy catheters is what I would have. Because I think
it's very difficult to do an effective arterial repair without the
ability to do the thrombectomy, proximal and distal, because you will be
repairing that segment into thrombosed vessel with thrombosed inflow and
thrombosed outflow.

I think you have to get that vascular repair kit, if you will, assembled
and communicated with your scrub team ahead of time, so as to include
the specifics of each item: specific heparin saline dose, shunts to be
used, SH needle and some of the smaller ones, and thrombectomy
catheters. It is important to go through that series of tools that you
may find yourself needing. Also, keep in mind, we don't use pulmonary
artery catheters as much anymore, but they can be used as a thrombectomy
catheter in a pinch. And then those thrombectomy catheters can be used
as good proximal inflow control. You can insert a large thrombectomy
catheter proximal and inflate the balloon with a three-way stopcock and
that will afford you inflow control in a pinch as well.

**Say you're having a patient transferred in from the Role II to the
higher level of care. Do you recommend these patients be heparinized, if
able, when they have shunts whether arterial or venous**

If they can be, then yes. It is dependent on whether there are other injuries to the head, torso, extensive soft tissue, etc. If there is massive bleeding, then the recommendation is to not shunt or put these patients on heparin. 

The best anticoagulant is flow. If you have high flow in those larger proximal vessels (above the knee or elbow), the shunts will stay patent without heparin (typically for 4-6 hours). The distal shunts, if they do clot off, what we have found is that they do not cause harm when they clot off. You
basically are just back to square one with the occluded vessel and you
have to do the thrombectomy, assessment, and repairs.

So even if the shunts do thrombose, our experience has been -- at least
in the short term of the temporary shunts -- that in two, four, five,
six hours they do not cause harm when they thrombose. You just have to
do the thrombectomy and assessment at that time.

**To close out iliac artery injuries, if you did need to ligate either
the common or external iliac artery due to severe injury, what kind of
options do you have there?**

Well, that all really depends on the patient and their overall
condition. If you have found yourself not being able to put a shunt in
and in that scenario you had to ligate the external iliac, that patients
is presumably in a really bad situation.

You can ligate it temporarily. So thinking that I am going to ligate it
and then get some help. But in 2-3 hours, I am going t o have to go back
and try to re-establish inline flow with another set of hands, with more
blood, a different surgeon, different lighting, different tools, because
leaving it ligated will incur a tremendous amount of ischemia into that
extremity.

There are the options of a cross femoral graft. I don't think I have
ever done that in the setting of trauma, bringing a PTFE graft from the
other groin, the other femoral artery over to the injured femoral
artery. That would be extraordinary or unusual. But when the options are
just life over limb, and you have to just say: "I can't repair it." The
scenario, whatever the reason, does not allow me to establish inline
flow and the patient's going to get an amputation and we are going to
work on saving his or her life, and move on.

**As we move our way down the leg, I want to stop and focus on the
common femoral and just discuss the principles of controlling the vessel
--- the options and if ligation is an option at this level for any of
the branches? Also, in residency, we would talk about blast injuries and
accessing the common femoral to get hemorrhage control. Is this a common
maneuver performed in the combat trauma?**

The common femoral is not much less consequential than the iliac. In
fact, it's probably every bit as consequential, as far as ligating it.
So it can be ligated as a damage control maneuver, but the consequences
will be significant if ligating the common femoral -- just as
significant as the external iliac.

There's three femoral arteries: the common, the deep femoral, and the
superficial femoral. I think similar to what I mentioned before, all
efforts should be maintained to try to maintain or establish flow
through those three, if at all possible, and balancing the situation at
hand, which sometimes means I have got to ligate it -- because of the
flow of casualties, because of the capability I have at a given leve.
Again, ligation if its life over limb is still an important damage
control maneuver. You just have to be ready for the consequences of that
and move on.

So controlling it, I think the Fogarty catheters are useful to control
the deep femoral. Sometimes from the open common, you can insert a small
Fogarty down into the orifice of the deep femoral and inflate that
balloon and control back bleeding from the profunda. Often there are two
profunda femoris arteries. They are difficult to control as they are
right adjacent to the deep femoral vein. So, oftentimes controlling
those deep femoral vessels from the inside using the small Fogarties and
a three-way stopcock is quite handy. Then, the same tenets apply with
either placing a shunt and trying to do a primary repair or some sort of
interposition repair.

**The popliteal artery is an area that we frequently see injuries and
they can be quite difficult to manage. Can you discuss just a little bit
about how you best expose and set up the patient when you are trying to
access the popliteal artery, both above and below the knee.**

Yeah, I will. And I skipped a question there. You asked about accessing
the common for control. I think that's acceptable, I mean, it just
depends on where the injury is and who is managing the case. I think if
it's an individual who has done, 15 or 20 or 30 of these sorts of cases,
and they feel comfortable about accessing the common femoral to control
it -- I don't think there's anything at all wrong with that. On the flip
side, if it's a distal SFA (superficial femoral artery) injury or a
popliteal artery injury, in those situations, most of the time it should
not require just a virgin cut down on an uninjured common femoral.

Now for popliteal, I think it is really important to get in-flow control
to the popliteal. One has to expose the above knee popliteal artery,
which is really the distal segment of the superficial femoral artery as
it comes through the adductor magnus or Hunter's canal.

You would start with a wide, long incision from above the knee to the
mid thigh. It is important to make sure to have a small sort of bump
with rolled towels that are placed underneath the calf, so that you prop
the leg up in a "frog leg position" and let gravity pull a lot of the
musculature of the leg down. Then, find the the distal superficial
femoral which is proximal to the above knee popliteal, just at or beyond
Hunter's canal.

And now move the bump above the knee. Then, make a separate medial
incision below the knee, again using sort of gravity in a way to pull
the gastrocs and soleus down, so that you can open up that below knee
popliteal space -- you are taking the muscles right down on the medial,
inferior edge of the tibia to open up that below the knee popliteal
space.

In this situation, you are going to need Weitlaner retractors.
Oftentimes, we have something called a popliteal retractor, which has
different depths of blades on the retractor. That's also referred to as
a Henly popliteal retractor or Pilling retractor (must be the company
that makes it). Those are awfully handy because you may need varying
depths in blades, deeper than the standard Weitlaner retractor.

And then being able to really put a narrow handheld retractor, like an
appendiceal or a Wylie renal vein retractor, that allows you to really
retract. If you are above the knee and you are not that well exposed,
you can put that narrow handheld on the inferior aspect of the wound and
sort of toe in and really get down into the popliteal space itself.

If you are below the knee, you put that narrow handheld in the proximal
extent and again, toe in and you really can try to get that popliteal
space itself exposed. If you need to, you can join the above and below
knee incisions -- sometimes with the trauma, the injuries are joined
anyway. Those are sort of some tenets.

Lastly, lighting. It can not be overstated how important lighting is,
sometimes that's a luxury for those of you who are in a deployed
setting. But these are the top line things to think about with popliteal
artery exposure.

\*\*From one of my own clinical experiences, especially in young
patients, with repairing or bypassing the popliteal artery, many times
these patients have very severe vasospasm at the end of the case. So
that, even though you may do a beautiful bypass, you have minimal
Doppler signals in the foot at the conclusion of the case. Have you
experienced this and do you have any thoughts on this?

Yeah, for sure. If the bypass has sort of an above knee-to-below knee
and you have excluded or ligated the injured popliteal artery, then
absolutely, a lot of times the patient will be relatively cold and in a
vasoconstrictive condition. As long as you feel like there's flow in the
reverse vein bypass -- ie: you can hear audible flow and it's not a
water hammer signal, meaning that there's some diastolic flow in the
arterial signal, and then there's some sort of weak signal at the ankle
-- most of the time, I will just be done, warm the patient up, and
resuscitate them. If any of those things are not true, you don't have
any arterial signal in the graft, or if you do and it's only a water
hammer signal, or where there's no signal at all at the foot or beyond
the graft -- then, you have to open it up. Pass thrombectomy catheters,
proximal and distal, and get that regional heparinized saline, and
really make sure that there's not a real problem. But, that's really a
judgment call and it is anxiety-provoking for sure.

But you have to have faith. I do think that if you feel like technically
you saw all of your stitches, you felt it was good, you had some good
outflow back bleeding before you did the distal, you didn't forget to
the thrombectomy proximally, and you have signals -- even if they are in
spasm and you feel it's not as good as it should be -- in some of those
cases, you warm the patient up and resuscitate them and the signal will
improve.

You can also do an on-the-table arteriogram, but that puts you in an
austere location and can be a little bit of going down a rabbit hole.
You will see the spasm and then say: "well, geez, now i need to
intervene on the spasm." An on-table arteriogram is certainly an option
and it should not be discounted, but I also would encourage folks that
as long as those things are true -- you saw your anastomosis, you
thrombectomized proximally and distally, you feel good about your
inflow/outflow, you have an arterial signal that has got some diastolic
flow (not water hammer in the graft), and you have some signal beyond it
-- warm up the patient and see how they do.

**As we continue down the leg, we encounter our tibial vessels. How do
you decide on repairing tibial vessels versus just ligating them?**

So there's a great paper I would refer you and your listeners to that
was published in the Journal of Vascular Surgery in 2010 with Gabriel
Burkhardt as the lead author.[@burkhardt2010] It was a review of the
U.S. military experience with the practice of selective tibial artery
repair. When you think about what that means, selective repair, it means
we repair some but not all tibial arteries. So how do you determine
which ones to repair? That was a good review article in the Journal of
Vascular Surgery that I think looked at maybe a hundred tibial artery
injuries. What we found from that experience was some principles of
selective repair which led us fix those patients in which all three
tibial arteries were injured. This means that it was not just the
posterior tibial artery, but because of the injury pattern, the peroneal
and anterior tibial had also been severed and there was just no flow in
the foot. So in those situations, we did repair them with a reverse
saphenous vein graft. That's the exception. Majority of times, as you
are alluding to and in that case series, we did not repair them because
we could ligate the single tibial artery injury since there was
redundant arterial flow to the foot. So you can ligate it, listen on the
table, and if the other tibial vessels are patent then it can remain
ligated and should remain ligated because tibial artery repairs are even
more technically challenging and time-consuming.

**We are going to shift our focus to cover venous injuries. What are
your thoughts? We have been talking a lot about the possibilities of
ligating arterial injuries. How do you approach venous injuries and
ligation?**

This would also fall under the descriptor of selective repair, which
means you repair some but not all. Without going into too much detail,
it depends. In almost all of the upper extremity veins, can and just
should be ligated. There is not enough muscle mass and flow in the upper
extremity veins to bother with repair. The exceptions might be a
proximal axillary vein or subclavian vein. But, by and large, the distal
axillary, brachial, basilic, cephalic veins can be ligated. In the lower
extremity, similarly, the small tibial veins can and should be ligated
-- there should be no attempt to repair those. Coming proximally in the
lower extremity, popliteal and then the femoral vein, if the patient is
physiologically stable and performing a repair is possible or placing a
venous shunt (we have had good luck venous shunts staying patent), then
we have deferred to fixing the veins in those cases. If the patient is
not in profound shock and if the expertise is there and present to do it
with operating time, I think there is benefit to fixing those veins.
There have been studies that confirm that this approach affords better
arterial flow in the arterial repair because the venous outflow is not
impeded.[@quan2008; @clouse2007; @rasmussen2022] Even if those veins
thrombose over time, they do it slowly and do not lead to pulmonary
emboli. So, when we can, we will fix those veins in the popliteal and
femoral segments. Conversely, if the capability does not exist, the
surgeon is not accustomed to repair, the patient is not doing well
physiologically, or there's the need for the operating table -- then you
just have to ligate those. I think ligating those veins in the popliteal
or femoral segment are also acceptable as a damage control maneuver.

**We could not finish a peripheral vascular trauma talk without
discussing fasciotomies. What are your advice for listeners regarding
fasciotomies?**

In the military setting, we have been liberal about the performance of
two incision-four compartment-lower extremity fasciotomy. I think they
should be done more liberally in our setting in the military. We
advocate for them and have shown that they can be done without
substantial additional morbidity and mortality. In fact, waiting to do a
fasciotomy or missing a fasciotomy has been shown to be associated with
mortality from early data from the wars. So part of that is in our
system, we often are going to lose track of these patients. We will see
them at a Role II, and then we would like to think, they don't need the
fasciotomy at the Role II because they will do it at the Role III. And
then we are not going to do it the Role III, because they can do it in
the Role IV. But again, we are going to be evacuating these patients at
altitude and they are going to go out of our care. So I err on the side
of prophylactic, two incision-four compartment fasciotomy in the setting
of lower extremity vascular injury certainly with wartime vascular
injuries.

## Abdominal Arterial

**22 Dec 2019:** *Kevin Kniery, MD, MPH; Adham Elmously, MD; Todd
Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Abdominal_Trauma_Part_One.mp3")

```

**We are going to skip the preoperative workup and description of
anatomy and zones of abdominal vascular trauma, so please review this on
your own. As a side note, we are also going to skip the management of
iliac artery injuries as we covered that as part of our peripheral
vascular trauma session.**

**Scenario: "You have a patient with a gunshot wound to the abdomen.
General surgery has already prepped them from the neck to the knees, and
performed a laparotomy from the xiphoid to the pubis. They packed the
abdomen and temporarily have control. There was note of pulsatile
bleeding from Zone I. Can you take us through how to get supraceliac
control?**

A case like this should not proceed without close communication of the
entire resuscitation team. I would start by making sure that I was on
good terms with my anesthesia colleagues. So, they know what we were
facing in this situation and that they were pursuing all the key tenets
of damage-control resuscitation --- leading with blood product or whole
blood resuscitation, keeping the patient warm, etc. As far as from a
technical standpoint, you can't sew what you can't see. So, I
immediately assess my exposure --- which is the laparotomy. The
laparotomy in this case was from the xiphoid to pubis. That's a great
start, but for the approach you have referenced, we will need to come up
higher and go between the costal margin and almost begin the first 2 to
3 centimeters of a sternotomy. It is not a sternotomy, but, it comes
along the side of the xiphoid to release the inferior costal margins of
the thoracic area under the diaphragm to pull the costal margins
laterally and up. You want to suspend your costal margins so that you
are almost suspending the patient's rib cage up and away from the aorta
and the abdominal contents. So making a good assessment of your exposure
initially is really important. In a case that I am prepping and
operating on, I try my best to use a retractor. Such as the Omni, that
will allow for retractors to be placed underneath the costal margin so
one can spread those costal margins and try to lift the thoracic area
away from the abdominal contents. If you do not have an Omni retractor,
you can still makes sure that you are suspending the superior aspect of
your exposure.

Once that is done, the other part of communicating with anesthesia is to
see if they have a nasogastric tube down. In this case, supraceliac
control is almost always facilitated by having a nasogastric tube in the
esophagus to allow one to open the esophagus here and be able to get
around that esophagus circumferentially --- with either umbilical tape
or penrose drain. This will allow for appropriate left-sided retraction
of the esophagus, away from the crus that is surrounding the aorta.

The next maneuver is mobilizing the left lateral segment of the liver,
so that it can either be brought cephalad or tucked inferiorly and
brought to the right side (or the whole portion of the liver). So now,
you have the left lateral segment out of the way. At this point, one is
left with the crus that is overlying the aorta.

The other maneuver here, dependent upon the situation, is to put the
patient in reverse Trendelenburg to let the stomach down and pull it
inferiorly so that you create as big a window as you can --- with the
esophagus retracted to the patient's left the stomach pulled down.
Sometimes, you can have a resident or a an assistant put their hand on
the stomach and retract it towards the pelvis.

Now, you have a window in which you can work. Next is to divide the
skeletal muscle, which is the crus overlying the aorta. This can be
facilitated with a large right angle and a bovie extender, and hopefully
you can feel the aorta with a pulse (even a weak pulse) and divide the
crus over the aorta with blunt dissection on either side. Typically, you
do not try to spend time to get circumferential control of the aorta. It
is not like an elective aneurysm repair where you want to get around the
entire aorta. In these cases, if you can get on either side of the aorta
down fine, that is what is needed and is most efficient in these
emergencies.

**Scenario: You enter a patient's abdomen and there is a large rush of
blood. You think it may be coming from somewhere in the upper aorta. Is
it worth taking the time to get supraceliac control before doing a
medial-visceral rotation and exposing the aorta in that way? Or are
there any scenarios in which you do a medial-visceral rotation first
before you get supraceliac control? Can you take us through a
medial-visceral rotation and pitfalls/advantages of doing each of those
approaches, ie when to leave the kidney down and when to take it up?**

It is worth getting super celiac control if it is a Zone I hematoma,
especially if you have some time with someone holding pressure. Now, if
you initially believe that the bleeding is from Zone I, but when you are
actually able to enter the abdomen it looks like the bleeding is coming
from the spleen or a different area, then you can readjust. The left
medial-visceral rotation is designed to give optimal exposure of the
para-visceral segment of the aorta. The para-visceral segment of the
aorta includes the supraceliac segment all the way down to the aortic
bifurcation, primarily the left common iliac artery. In this case with
the left medial-visceral rotation, it is performed through dividing the
white line of Toldt along the sigmoid colon and the left colon. You want
to establish this plane in the retroperitonium, which a good portion is
achieved through steady blunt dissection facilitated with Bovie
electrocautery. In the setting of trauma, this dissection plane is taken
from the sigmoid/left colon up to divide the attachments ingrained in
the spleen and the left kidney, then to the midline and then ultimately
to the patient's right. In the setting of trauma, the left kidney is
almost always brought up. It is generally easier and faster, because you
do not have to take the time to find the plane above the left kidney.

Check out this [video]( https://www.youtube.com/watch?v=DPxiDYzJwcY) for an example of a left medial visceral rotation

**Management of bleeding from the spleen**

One of the key tenets of splenic injury is exposure --- good retraction with incision up to xyphoid and
alongside the left costal margin. Compromised exposure is a contributor to splenic injury. 


**With regard to exposure and eviscerating the bowel, there are some
nuances. A method includes finding a moist lap pad that is large enough
to cover the bowel, then place a clamp around it with someone holding it
outside the abdomen. However, this method may not be as secure and also
result in the bowel still being in the way. Do you have any tips on how
to eviscerate and control the small bowel?**

This is difficult to describe, because sometimes you just have to feel
it. However, this is the main purview and job of the surgeon that is on
the patient's right. It takes a strong and big right hand, extending as
wide as possible, to get the bowel over to the patient's right side ---
pulling it from the left to the right. It is not a subtle maneuver.
Also, I do not use lap pads. I use an operative towel and ask to moisten
the blue towel. You take the blue towel in your left hand and put it
underneath your right hand (which has the viscera) to try to get it down
to the base and laid out --- the blue towel and your right fingers are
essentially coming across the aorta passed midline. Then, my preference
is to use the Omni retractor to take the place of your right hand, which
will come on top of that blue towel and hold all the viscera to the
right of patient's midline.

**With the left medial visceral rotation (Mattox maneuver), we perform
the maneuver because of a suspected aortic injury or a close branch off
the aorta. However, sometimes we need to do a right medial visceral
rotation (Cattell-Braasch maneuver), especially when concerned about an
IVC injury. What are some of the nuances of a right medial visceral
rotation?**

As you are on the other side, it still requires dividing the white line
of Toldt starting at the cecum and coming along the right colon. You
bring the cecum and right colon cephalad and then to the patient's left.
It is actually not so much of a rotation to the midline and then the
left, but bringing everything up cephalad and then left to the patient's
left costal margin --- almost bringing the cecum and right colon up onto
the patient's left chest. This is a plane that is also similar to the
medial visceral rotation on the left side as it is largely developed
with blunt dissection. However, it is a combination with opening the
retroperitoneum with some Bovie electrocautery. You will also almost
immediately see the left iliac vein and then the vena cava. Then, you
continue to bring the left colon up until you see the right renal vein
and kidney. At the superior extent, this will expose the head of the
pancreas and duodenum. And similarly, you use a blue operative towel to
cover and bring up the viscera to be held with the large retractors so
your hands are free to operate. Spine exposures with orthopedic and
neurosurgical colleagues and transplant procurements are also a great
opportunity to learn the tenets of rotating the visera either right or
left and exposing the retroperitoneum. 

Check out this [video](https://www.youtube.com/watch?v=KVuMiAuw5zc) for an example of a right medial visceral rotation.

**The technique for exposure of the infrarenal aorta**

The exposure of the infrarenal aorta is also called the transperitoneal
inframesocolic exposure and is considered to be a straightforward
manevuer. Inframesocolic means that the transverse colon is reflected
cephalad. One of the first aspects of this maneuver is to hold up the
transverse colon and drape it up cephalad to patient's costal margin
(can be held up with a blue towel and retractor). Next, you divide the
ligament of Treitz (by the surgeon on the patient's right) and get down
to the fourth portion of the duodenum which should be to the left of
patient's midline. Once divided, the left colon remains on the patient's
left and everything else goes to patient's right. However, this is not a
medial visceral rotation because only the fourth portion of the duodenum
and all of the small bowel is partitioned to patient's right. At this
point, one should see the infrarenal aorta. The left renal vein is the
most cephalad extent and the aortic bifurcation is the most caudal
extent (located at root of colon mesentery). The left renal vein can be
ligated if necessary to facilitate aortic control. However, it is best
to try to preserve the branches of the left renal vein that drain the
left kidney (gonandal and lumbar) so that dividing the left renal vein
is not as much of a detriment to the kidney.

**Tips for getting umbilical tape or vessel loop around aorta,
especially in supraceliac region.**

You can achieve this with a combination of Metzenbaum scissors and
controlled pressure. You are trying to free the aorta from the
connective tissue around it to get back to the spine. You can use a
metallic pediatric yankaeur (small profile and blunt tip) as a
dissection tool and push with the Metzenbaum scissors along with direct
feel to get the aorta freed. Then, you can use a right angle to get
underneath the aorta. You also want to make sure to identify the lumbar
vessels --- "find the outside before you find the inside" --- so you can
clip them with medium/large Weck clips.

**Techniques of exposing and repairing branches off of aorta: Celiac
Artery**

The origin of the celiac artery can be seen with left medial visceral
rotation as it will show the entire paravisceral segment of the aorta
and also with an anterior approach through the lesser sac (very similar
to exposure needed for the supraceliac aortic exposure). The stomach is
retracted caudally towards the pelvis, lesser sac is opened, and then
the stomach is pulled down further. The pancreas is pulled down as well
and the origin of the first centimeter of the celiac artery is now
exposed. For injuries to the proximal portions of the celiac, if able to
be repaired primarily --- use 4-0 or 5-0 prolene sutures. If injury is
large and there is destruction of vessel wall, then primary repair is
not an option and this becomes a damage-control situation to control the
bleeding. With damage control, ligation is even considered as an initial
step to save a patient's life.

**Techniques of exposing and repairing branches off of aorta: Common
Hepatic Artery**

The common hepatic artery is best visualized through that similar
transperitoneal anterior approach for exposure of the celiac artery
through the lesser sac. Prior to identifying the common hepatic, you
have to find the celiac and follow the branches. Patient should be in
reverse Trendelenberg, the stomach inferior, and costal margins flared
up to have the appropriate space. This will expose the second and third
portions of the celiac artery, and its major branches --- including the
common hepatic. The common hepatic artery can not be visualized using
left medial visceral rotation. Primary repair is ideal, however, in
damage control surgery ligation may be the only choice. If possible,
ligation of the common hepatic proximal to the GDA is preferrable as
there will be retrograde flow from the GDA into the proper hepatic to
maintain the viability of the duodenum, pancreas, and liver.

**Techniques of exposing and repairing branches off of aorta:
Gastroduodenal Artery (GDA)**

The gastroduodenal artery is the first branch off of the common hepatic
artery and defines the junction between the common and proper hepatic
arteries. Primary repair should be sought if possible. If it is a damage
control situation, then the GDA must be ligated to control hemorrhage.
If ligated, assess patient's condition, resuscitate, utilize the
doppler, and assess the area distal to ligation to evaluate the
consequences. For example, if ligation showed appropriate collaterals,
then the vessel can remain ligated. However, if there is clear ischemia,
then there is a difficult decision between repair vs. shunt vs.
permanent ligation along with its metabolic consequences. The decision
is dependent on each situation. The principles of damage control
resuscitation and damage control surgery should always be maintained.

**Techniques of exposing and repairing branches off of aorta: Superior
Mesenteric Artery (SMA)**

The SMA is divided into segments based on its course, from aorta to
pancreas to root of mesentery to the small bowel. Approach to SMA
visualization is dependent on the segment that needs to be exposed. The
first segments of the SMA can be seen through left medial visceral
rotation. The distal segments of the SMA can be seen through the
anterior transperitoneal approach. Since the SMA (second segment) is
near the head of the pancreas, if the proximal segment of the SMA is
injured, then the head of the pancreas may need to be divided to get
appropriate exposure. If the segment of the SMA that is injured is
distal to the pancreas, then it can be visualized through the
inframesocolic approach. With regards to ligation, distal SMA segments
will be more tolerant of ligation. All efforts should be made to
mantatin flow through the SMA. Primary repair can be performed on small
injuries with 4-0 or 5-0 prolene suturing. Larger injuries can be
repaired with patch angioplasty or interposition graft (PTFE, Dacron, or
autologous tissue such as proximal greater saphenous vein or deep
femoral vein). Autologous vein is preferred in contaminated field, but
can also use rifampin-soaked Dacron. Tunneling is dependent on location,
proximal segments would need 6-8 mm of graft while distal segments would
need 3-4 mm. It is important to remember that a majority of these axial
injuries are typically unsurvivable and proximal SMA repair is uncommon
in the trauma setting. Even if the SMA is repaired and the bowel appears
viable, it is preferred to delay abdominal closure and perform a second
look 24-36 hours later or sooner depending on the patient's condition.

**Techniques of exposing and repairing branches off of aorta: Inferior
Mesenteric Artery (IMA)**

The majority of times, the IMA can be ligated. Prior to ligation, assess
the patency of the iliac vessels --- specifically the internal iliac
vessels. Also, if both celiac and SMA are patent, then the IMA ligation
can be tolerated. If there is concern for atherosclerotic disease or
occlusion in the internal iliacs, then ligation may not be the
prefferred option and re-implantation may be considered.

**Techniques of exposing and repairing branches off of aorta: Renal
Arteries**

It is important to consider the broad principles and approach: 
1. Urgency of situation 
2. Optimal Exposure 
3. Resuscitation Efforts
4. Communication with Team
5. Appropriate Equipment 

Decision to maintain patency of a vessel vs. ligation is dependent upon the situation. During damage-control procedures where bleeding is not controlled and location of hemorrhage not known, the inclination is to ligate. The risk is organ ischemia, but the benefit would be patient survival. 

The renal artery is similar to the SMA, unlikely to be reconstructed in a trauma setting. In renal trauma, the renal artery is either not injured (with only injury to renal parenchyma) or renal artery is injured and thrombosed. If renal artery is injured and thrombosed, salvageability is unlikely due to prolonged warm ischemic time of the kidney. Repair of the renal artery would truly occur if there is an expanding hematoma in Zone II that warranted exploration and it was found that the source of the hemorrhage was from an injured renal artery that still has flow. In this case, again, would do primary repair with smaller injuries with a 4-0, 5-0, or even 6-0 monofilament prolene suture. If a primary repair would compromise the lumen, then consider patch angioplasty. However, most commonly, if renal artery is injured it is ligated for damage-control. With regard to incidental CT findings, an intimal flap , fistula, and/or possible pseudoaneurysm of the renal artery, the injury can be observed if there is continued renal perfusion and no flow limitation. If it is a flow limiting injury with more than 50% stenosis caused by the intimal flap, it may be amenable to endovascular placement of bare metal stent to tack down the flap. Distal branch artery injuries can also be managed with endovascular coil embolization. Again, the decision to take the time to coil embolize is dependent on the patients current hemodynamics. Most commonly, if there is an expanding hematoma in Zone II and the renal artery/parenchyma is damaged and confirmed as source of hemorrhage, a nephrectomy is warranted and there is no indication for repair. 

**General Guidelines of Penetrating vs. Blunt Retroperitoneal Exploration** 

Zone I hematomas: explore whether penetrating or blunt mechanism.

Zone II hematomas: always explore if penetrating, do not explore during blunt unless expanding. Kidneys are in Zone II and are in the lateral gutters. 

Zone III hematomas: always explore if penetrating, do not explore during blunt unless expanding.


## Abdominal Venous

**18 Jan 2020**: *Kevin Kniery, MD, MPH; Adham Elmously, MD; Todd
Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Vascular_Venous_Auphonic.mp3")

```

# Trauma - Abdominal Venous

**Kevin:** 

Okay and welcome back to the behind the knife and audible bleeding vascular trauma podcast series with our guest Dr. Todd Rasmussen Dr. Rasmussen is a Colonel in the United States air force and is a professor of surgery and associate Dean of research at the uniform services university of health sciences in intending vascular surgeon at Walter Reed military medical center thank you Dr. Rasmussen for coming back on a joint podcast to help us learn about trauma 

**Rasmussen:** Sure it's a pleasure and I look forward to some of our discussion Thank you 

**Kevin:** We're lucky enough to have one of my partners dr OD Hum A L muesli He's a chief resident at Cornell and he'll be helping us with this interview Thanks Adam

**Adham:** Thank you I'm happy to nerd out about some surgical technique Was a master here 

**Kevin:** So We're going to drive straight into abdominal venous injuries and we're going to keep it simple here When we're thinking of abdominal venous injuries we're gonna think of three types You have your IVC you have your portal vein and you have her superior mesenteric vein So we're going to break these down in order and discuss the approaches to exposing it and fixing it So of these doctor recipes and what what do you see most commonly 

**Rasmussen:** that's a great question is to which we see most commonly I think they're all relatively uncommon because of the frequency of blunt trauma especially in the civilian setting I suspect that blunt liver injury and the relationship between the liver and vena cava 

**Kevin:** In our isolated injuries to the the IBC is there any non operative management that's a minimal for the IVC 

so I think yeah I think there's a number of types of injuries Scenarios would be amenable to non operative management we just mentioned very briefly the retro Paddick cable injuries in fact most of the time if the patient and be stable and doesn't have other associated injuries that require exploring that area of the vena cava Oftentimes retro hepatic vena cave injury should be managed non-operative And that's of course the difficult nature of exposing that area of the of the VTK VA and the high risk of causing more bleeding with with that exposure because of the venous system the vena cable has a low pressure system infer renal vena cava injuries super Reno vena cave injuries or even super paddock can be well tolerated the nonoperative management cause it's a low low pressure system but in particular the retro retro Paddick a portion I think is 

**Adham:** But you kind of touched on this a little bit but when when we talk about diversity We sort of divided into four segments anatomically because each of them has their own implications and difficulties The infer renal segment the super renal segment and then the retro paddock and through prophetic segments or the vena cava I think it's probably worth going through each of those segments And Discussing briefly exposure and I think we should spend some more time on retro hepatic but discussing exposure and then strategies for repair versus ligation and their implications in each of these segments 

**Rasmussen:** Okay Yeah So I guess we'll start with the infer renal segment of the vena cava It's probably the most easy to expose this the infer renal Being a cable segment is best approached through a right visceral rotation meaning of bringing the right colon the and the viscera up cephalad in the retroperitoneal space to expose well to expose the vena cave in the infernal segment And That maneuver at its proximal most extent will will expose the the renal veins the fourth portion of the Dudina and the pancreatic head so that's that's the best way to expose that and then I think repair options there it depends upon the severity of the injury if it is A partial injury to the vena cava then hopefully primary repair you know the vena cave is quite large it has a large capacity and controlling that and and and and repairing an injured segment primarily with with a four O proline a suture example is pretty well tolerated If one feels like they're going to narrow the vena cava significantly than that's a different Consideration which includes then should the BTK would just be like as opposed to trying to repair the vena Cavan and leave it narrowed which then would predispose it to a thrombosis and potentially a life-threatening thromboembolism So that's sort of this infer renal segment I mean the I mean we could talk an hour on this segment in the vena cava Controlling is challenging because there's it bleeds black blood and in large quantities and oftentimes controlling proximately distance is not enough because there are lumbar vessels the lumbar veins that are back behind that if you know one doesn't control one of the large lumbar vessels that's coming into the vena cava Posterially you can have all the proximal and distal control that you want and you still can't see anything because you've missed a huge lumbar vessel That's you know lumbar vein that's feeding from the backside so one has to be mindful of that and try to either identify that and get it with with a clamp of some sort or get around it with a vessel loop or something to control it as we have said in other of our discussions none of these operations or these techniques take place in isolation and they've gotta be done in close communication with anesthesia and assure that anesthesia is not only Adhering to the principles of damage control resuscitation with you but they're also aware of these maneuvers whether it's losing large volumes of black blood from the vena cava or maybe like gating the vena cave and the infernal segment which is going to significantly reduce the inflow to the right heart 

**Adham:** So two followup questions on the infer renal cable in general you said you use a O pulling to repair the Kiva Do you ever Pacific needle Probably a larger needle is better And then the second question is when you're doing and for renal cable ligation do you think we should be empirically doing a four compartment Fasciotomy is on these patients 

**Rasmussen:** Yeah So really good point about the needle size So I think a four O on an sh needle is imperative So these are the subtle parts that I think before someone tackles this needs to have in mind Generally you need to barge for suckers You know you've got to have headlights and extra set of hands because you know you may only have one or two shots at this you know controlling it and seeing where the bleeding's coming from And and to the best you can don't waste one of those opportunities cause what you're going to be able to do is control it with for example sponge or even your hand so you know while you've got a controlled then make sure you've got the right lights the right suction devices the right assistance blood in the room you know all of those things Take a deep breath and get all that squared away To your point of a needle the point here that we're talking about for your listeners is don't use too small of a needle a common mistake is a this has gotta be a dainty needle It's gotta be small because there's sewing on the BTK BA may be a Five-O or you know six O proline and I'll put it on an RB one or C one needle These are small needles and what happens is that needle You can't see it in all of the blood So an sh needle is really designed with a sort of a large arc that really can literally go underneath your finger and a four O proline will go on an sh needle So absolutely as far as the empiric performance of four compartment fash autonomies with ligation of the Munich kava I'm going to take a pass on that and say it depends I think it depends upon how much resuscitation the patients received meaning if they've received large volumes of fluid they're physiologically not well and you know you anticipate large sort of significant recovery burden then Yes I would probably do of fasciotomy is of the lower extremities and elevate those legs and do everything you can to keep the pressure down In contrast if the patient's physiologically pretty well that was a relatively small resuscitation then you know perhaps not I think it it just depends but you should be definitely mindful of it there's no doubt And and if you don't do the fast shot dummies recognize that she got to keep an eye on the leg and the legs elevator 

**Kevin:** So a doctor asked me he and I have to thank you we actually kind of talked about managing IVC injuries in our first episode So there's this is our third episode total We've covered lower extremity injuries We've covered abdominal aortic vascular trauma in today's our Venus But on our first episode we talked about using the large FORO needle and about a week after that episode we got called into a internal hemipelvectomy case which I didn't even know it was a case where the orthopedic surgeons were resecting this massive sarcoma off the pelvis and basically resecting half of the pelvis and they got into massive bleeding Basically they had gotten into the crotch of the IVC and had just extenuating hemorrhage near move their hands and using the large Mauro needle The large needle in the Fort with is really helpful A because like you said it how did you see it through all the bleeding And then the thing is nice about the FORO is you can really kind of crank down on it and and push You're not down there without worrying about it breaking And when you're in a in this case with lots of bleeding and there's five people around the operating table I found it really helpful so when we were in that case 

**Rasmussen:** Good 

**Adham:** You mentioned in your book chapter which I also have seen in some liver cases were injuries So that gave are made That's helpful is when you're placing these sutures into the vena cava if you're doing a figure of eight or a figure of 16 placing the first suture and then lifting on both of those strands of the suture up in the air to elevate the hole 

**Rasmussen:** Yeah that's a an important part And I think you know I would go back to the fact that you you know you're only gonna get one or two maybe three chances at this So you know don't skive I mean don't be afraid to take too big a bite I mean you gotta get substance of the vessel and you got to get the bleeding control or you're just not going to to get to the next step So yeah So making that first pass that you think well this can be a figure vein it's a little bit blind It's not completely I'm not advocating for taking hail Mary you know I'm not saying that but But the point is be definitive about your suture placement and and go ahead and get full thickness by to the vessel wall then pull up on that and and that may help you expose you know each each injury is a little bit different but try to recognize that in order to get through it to that next step you're going to have to get a false full thickness bites of of even the vein you know and it in fact especially the vein right Because it's going to be more prone to tear So you really can't skimp on 

And so once you go above the renal veins the kind of the ability to to Leggett completely changes and everywhere above that So can we talk a little bit about the anatomy of the super renal and then just briefly discuss retro paddock and super hepatic IVC 

**Rasmussen:** Right So I guess we didn't really mention But the reason that the infer renal ligation can be tolerated it's not ideal but it patients will survive with ligating The Infor renal is because one has not disrupted the The venous outflow of the kidneys so with an infer renal ligation flow through the left and right renal veins is maintained and therefore kidney function is still preserved that has you referenced changes when the injury is above the renal veins in the super renal segment of the IBC and in those cases you're right ligation really isn't an option You know this is a very small segment as only a two or three centimeter segment of the IVC you know between the superior aspect of the renal vein and then the initiation of the drainage of the paddock veins you know the retro paddock So this is a very small segment and you know I think that if there's an injury in this segment it should be you know attempts should be made to primarily repair it if if you find yourself operating on it this also gets in this area where It really could also be almost considered retro Padick and certainly if these are blunt injuries or from a blunt mechanism you may not want to find yourself operating on that if the patient will give you that option 

so if you do find yourself needing to operate on this obviously you you need to mobilize the liver to expose this part of the of the vena cava what considerations do you need to take into account before mobilizing the liver And can you walk us through your technique in doing so 

**Rasmussen:** Yeah I think effective mobilization livered cannot be overstated how important that is because to to get enough control of the vena cave And this segment really requires dividing the attachments of the liver to the diaphragm and and getting the liver well mobile so that so that it can be moved Cephalad And allow exposure of this segment So I think what to be mindful of is is in this area this this is just a very difficult area because inevitably there are lumbar veins coming from the backside of of this aspect of the vena cava in inevitably to to control The the cephalad portion of this area there's likely to be a couple of hepatic veins that are coming into this area So it is in a mobilization liver I think is very important to to allow exposure I think we mentioned I guess to start the the medial visceral rotation of the right colon and the head of the pancreas all of that has to really be I think maybe to your point make sure that that exposure is also very thorough so that the the third fourth portion of the Dudina M the head of the pancreas are lifted up off of this segment so that one can really expose and see as we said the outset it's difficult to so what you can't see you know 

**Kevin:** And I think one point that we haven't emphasized enough here that you emphasize in your book substantially is the fact about really being careful about your decision to mobilize the liver and how you can really make your situation worse Can you touch on that 

There are some injuries where actually leaving the liver ACE of course retro hepatic If you get into the retro hepatic segment then and those situations often leaving the liver in places is I because it's it's What hemostasis may be attainable is unroof the tear and the vena cave and behind the liver So in those cases actually leaving the liver and not smoke might be beneficial and should be pursued in contrast if this is Injury to the to the super renal segment of the IBC It's a longitudinal tear and and you sort of see it that it's there It's not a retro Paddick segment then and you see the bleeding is from below the liver in the Vionic Gava Then those situations in mobilizing the liver to a degree is actually maybe necessary so that you can control of the injury the fatty grains the super hepatic Vena cava is that is that before the portions are we've talked about the infra renal super renal retro Patrick and then super panic that is exposing that There's also a pretty short segment air from the last of the phrenic veins that train into the new Kiva right at the cruise and then the writing heart and the pericardium So that's a pretty short segment there to expose that area one needs to really divide up along its IFAD to really get the costal margins to open up To just go from a tip of the to the pubis Often it is not enough Really in my experience takes a very good effort to to divide up along the the edge of the two to three centimeters cephalad separating the XY from the right costume origin to really open that up and show the all of the diaphragm and and at the base of that of course Will be the vena cava And that's sort of that fourth segment if you will that We have discussed 

Dr. Rasmussen at least in a lot of my training I've been taught with a large bleeding IVC to get proximal distal control sponge sticks you're in your book you have some precautions that you mentioned against this technique and how do you recommend approaching a 

**Rasmussen:** So I think the tenant of the sponge stick is is Is the same where the ease the sponge stick or and or a finger or another instrument you know it's it's I think it's the starter right You're trying to I got to start somewhere and there's this black blood welling up I've got pressure on it so that pressure can be applied with a sponge stick or a hand I think what one recognizes that whatever you use for that initial application of pressure you're going to have to refine your dissection You're going to have to refine your exposure and then you're going to have to refine that manual pressure mechanism because you can't operate everybody's got their hand in the woman it may not be bleeding but you can't be there forever and you can't see to actually disect and expose and operate So I think the the sponge sticks are fine I think so we don't skip over it if if this is a retroparotoneil hematoma from the inferior vena cava this has gotta be exposed in my experience through visceral rotation of the right colon the seek them and the right colon up to the patient's left exposing the vena cava You won't be able to manage this hematoma by going through the mesentery of the right colon for example That's so that's a premise that this injury has gotta be managed with a medial visceral rotation of the right colon and seek them a cephalad into the patient's left to expose this Injury Even if you start out with with the sponge stick or manual pressure before that maneuver let anesthesia catch up give blood in the room but eventually you're going to have to mobilize that seek them and right colon up and to the patient's left you know I think from there you know these technical tricks that you're referring to I think trying to get the application to something smaller than the sponge sticks So oftentimes something that's referred to as a Kitner or a peanut if you will those are lower profile you can see around them off and you can sew around them They're smaller than the deck and oftentimes use of those which is really the similar principle of Manual pressure is I think useful you know I think all you have the sponge sticks in to the degree that one can get proximal and distal control of some of the large tributaries feeding the injury area the iliac vein then that may be useful as well the other technical tenants are you're going to need probably two large bore suction devices lights blood in the room Don't overlook those things as we have Sat in a couple of our other segments you know this is the type of injury where you may only get one or two chances at this You want to make those those chances as good as possible So you know the the sponge stakes are fine but they're only going to be a starter and you're going to have to refine your your manual pressure to something smaller so that she can disect and expose 

so we were talking about retro paddock cable injuries and a technique that we didn't discuss that may be useful in situations where There was blood welling up from behind the liver and it's not clear where it's coming from is a total vascular isolation or total Astro exclusion you walk us through that 

**Rasmussen:** Yeah I took a good point And I think we haven't mentioned that And of course that is trying to isolate all the flow to the liver and reduce bleeding ultimately from the cave and the confluence of the of the paddock beans So you know if one thinks about the inflow into the liver right So that it would be the portal vein which is Probably 70 to 80% of the total flow So trying to isolate the portal vein and being able to control at least an isolate the poor to happiness and a portal vein the paddock artery through what's called a Pringle maneuver so that's the that's really the inflow to the liver and then as far as that segment of the vena cava controlling You know the vena cave and the renal veins below that injured segment the the other portion of inflow So going through those steps to isolate what is inflow into the liver and that injured segment of cave at where the confluence of the paddock beans is that's sort of the the key tributaries that need to be accounted for in that maneuver 

**Adham:** And then when you're above the injury and the super impatent Kiva Are you getting control in the area of the California so the paddock veins below the diaphragm or do you find that you need to go into the chest cave and the chest 

**Rasmussen:** Yeah So that's a great question I think when doing total hepatic isolation the the one segment which is a super hepatic segment of the vena cava I tend to cheat The question was do you go into the chest to control that segment or can you do it through the abdomen I always do it through the abdomen but when I say I cheat I suspect by the time that I I've done it I've divided the up along the cost of margin and almost started what amounts to opening the chest through the abdomen Because you're you're exactly right You have to get pretty high And using some deep what are referred to as long renal vein retractors the Wylie renal vein retractors that are handheld those can go down onto the cruise since it covers the vena gaven and you're basically in the chest or just in the chest through the superior aspect of the of the abdomen I can't emphasize in those situations how important it is to Of released the hepatic attachments to that portion of the diaphragm and then really are those costs or margins you know the cost of margins they tend to enclose around the abdomen and that's what they do But by dividing the three to four centimeters cephalad It's often overlooked but that really allows you to almost suspend the patient with retractors under the cost of margins And that really in a way puts you in the chest although you're there through the peritoneal cavity 

**Kevin:** And just for completeness sake to totally cover a total hepatic vascular exclusion if you still have bleeding after getting the super Padick and infer hepatic control and the Pringle is there an option to prevent any further flow I'm alluding to the point in your book where you talk about in this very rare situation you can do a super celiac aortic clamp to help identify the injury but the patients are unlikely to survive this 

**Rasmussen:** Correct I mean I think there's a couple of there's a couple of damage control steps You can certainly clamp the aorta to see if that helps I mean I think that's absolutely a great point and I didn't mean to overlook that if nothing else oftentimes clamping the is neededto support left ventricular afterload for a period of time because these patients are in rough shape or you may need to clamp the aorta as a sensitive maneuver anyway and they see that that helps or doesn't help the other The other situation is trying to then place that then bypass you know from the basically from the inferior vena cava up into the right atrium you know these are very rare but that is a way to then bypass the liver and allow a continued Right hard from the infer renal segment is by using a shunt So that has been described as the Strock shunt and most commonly is described using the chest tube and that can be used as well as another sort of last I don't want to say last last ditch effort but at that point like clamping the aorta you know you're pretty well down on the list of things that you are are are trying to do 

I think it is still a viable option I mean it's extremely rare but I would advise all surgeons to to learn about it and then to understand what it is I think it's unlikely that we'll find ourselves in those situations but any of the shunting techniques the temporary SA I think it's important to familiarize oneself with the with the principals and and to have in mind the situation in which which it might be useful So I don't think it's it's not any more condemned to the history books as any other you know a modality of temporary avascular shunting 

**Kevin:** Great And so as we're starting to wrap up our venous episode we've covered the four segments of the vena cava And just to recap that's their Infor renal which we can repair we can enter position we can leg gait if needed And then we have our super renal retro Pedic and super hepatic which are quite challenging injuries which we discussed in detail on kind of bail out maneuvers For those Now we're going to talk about our portal vein which also includes our superior mesenteric vein and options in this situation so obviously if possible we're going to you know repair an injury And obviously like we've said many times these injuries do not happen in isolation There's probably multiple other injuries to the pancreas and celiac artery and who knows what else but Can we just discuss If you have a portal vein injury you're probably in a pretty catastrophic situation 

**Rasmussen:** I thinkthat it's going to be tough So I think I mean I suppose they can tolerate it and like any damage control vascular scenario ligation is is an option you know and it may be necessary but I think ligating the main portal vein carries with it a very high mortality So I would if if I would consider gosh you know temporary vascular shontayne if at all possible although that's obviously pretty rare if it's like aided initially as a damage control maneuver is just trying to get things stopped then I would Get the patient stated and then I would try to reconstruct it you know because I don't think longterm the ligated gated portal vein is going to it's very unlikely to be survivable So if for example let's lie gait it let's get the bleeding stop Let's get a Another set of hands in here Okay we've got the bleeding stopped It looks like the portal veins now but let's get the patient resuscitated a little bit warmed up And let's 

And you mentioned in your chapter about you know for these injuries it's very likely you may need to divide the pancreas if the projectile As in divided it for you to help get better exposure to repair One thing I felt was really interesting in your chapter is you discussed how there's very limited experience with these injuries and and not a lot of followup but the thought is that with our current anesthesia techniques and our open abdomen techniques that these Injuries may be survivable when they weren't in the past can you talk a little bit about the blood return from the splanchnic circulation and and why it's important to discuss out the anesthesia and options if you do have to do an aggressive ligation and repair of this vessel 

**Rasmussen:** Well I think if I understand your question I mean as we've These torso vascular injuries are the most complex And and as we have discussed throughout our talk about vascular trauma the reconstructive technique whether it's a damage control technique or a definitive repair technique needs to be done in conjunction with Anesthesia and the resuscitation team and that's to assure that that that the tenants of damage control resuscitation are are be pursued by the team meaning blood and blood components placement prevention by Galapagos the mitigation of acidosis Optimization of relation all of those things So the I think the more complex the vascular injury the greater the consequences with that vascular injury the more important it is that we're communicating with the anesthesia team to let them know what we're doing And considering with regards to in this case these whether it's a portal vein injury or a superior mesenteric vein injury it's it's it's important So that damage control resuscitation occurs with damage control surgery one without the going to fail If you do perfect damage control surgery but you failed at the tenants of DCR damage control resuscitation 

And I so just to wrap up the the portal vein and we kind of Looping in the superior mesenteric vein into this Ligation is really not tolerated but in a catastrophic situation you can consider it as long as there's communication with anesthesia that 50% of the blood volume is going to be sequestered in the splanchnic circulation And then with management of open abdomens and aggressive fluid resuscitation there have been case reports of survival after portal vein ligation so Dr. Rasmussen once again We took you through a lot of vascular trauma today and you were very patient with us and breaking every detail down We really appreciate this and we think our listeners will also 

**Adham:** Thank you 

**Rasmussen:** Well I appreciate the opportunity and I commend you for taking such a complicated topic and I hope that it's useful and would encourage you know as we've said throughout our discussion encourage your Listeners to get an Atlas or a video and augment our discussion you know with with other learning techniques 



Check out Dr. Rasmussen's book

<https://www.elsevier.com/books/richs-vascular-trauma/9781455712618>

<https://www.amazon.com/Richs-Vascular-Trauma-Todd-Rasmussen/dp/1455712612>




## Cerebrovascular Trauma

**25 May 2020**: *Kevin Kniery, MD, MPH; Adam Johnson, MD, MPH; Nicole
Rich, MD, MPH; Todd Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Cerbrovascular_Trauma_auphonic.mp3")

```
**Neck Exposure Techniques**

Carotid Artery 
-	Anatomy 
        -	Right common carotid  originates from brachiocephalic artery 
                -	External landmark: right sternoclavicular joint
        -	Left common carotid  originates from aortic arch in superior mediastinum
        -	Common carotid artery (CCA) is within the carotid sheath 
              -	Contents: 
                  -	Common and Internal carotid (medially)
                  -	Internal jugular vein (laterally) 
                  -	Vagus nerve (posteriorly)
              -	Anterior margin: sternocleidomastoid (SCM) and omohyoid muscle
              -	Posterior margin: longus colli and longus capitis muscles
              -	Medial margin: esophagus/trachea
        -	Common carotid bifurcates to external and internal at level of superior border of thyroid cartilage
               -	External carotid artery 
                  -	Medial to internal carotid artery (majority of course)
                  -	First branch is superior thyroid artery (near bifurcation) 
                  -	Terminate in parotid gland, divide into superificial temporal and maxillary arteries
              -	Internal carotid artery 
                  -	No extracranial branches
                  -	Lateral to external carotid artery until level of skull base, where it crosses medially
                  -	Enters the skull base through the carotid canal behind the styloid process
              -	Facial vein is the anatomical landmark that approximates the location of the carotid bifurcation (deep to it). 
              -	Hypoglossal nerve (CN XII) and posterior belly of digastric muscle superficially cross the internal and external carotid arteries at the angle of the mandible 
              -	Glossopharyngeal nerve (CN IX) crosses in front of internal carotid artery, superior to hypoglossal nerve (CN XII) 
-	Exposure 
        -	Longitudinal incision along anterior border of SCM, extending from suprasternal notch to mastoid process
              -	For proximal common carotid artery control, would do combination of SCM incision with median sternotomy
        -	Incise platysma and expose anterior border of SCM, retract laterally
              -	Identify and avoid Accessory nerve (CN XI) that enters SCM
              -	Small branches of external carotid should be ligated to adequately mobilize the SCM and expose carotid sheath
        -	Carotid sheath is visualized and incised longitudinally
              -	Omohyoid muscle may require division if proximal exposure required 
        -	Identify Internal jugular vein, mobilize, and retract laterally
        -	Identify Vagus Nerve and avoid injury (posterior between internal jugular vein and common carotid)
        -	Place vessel loops around CCA, internal jugular, and vagus nerve
        -	Identify facial vein (overlying CCA bifurcation), Hypoglossal nerve (CN XII) and Ansa cervicalis 
              -	Ligate facial vein to further mobilize the internal jugular vein laterally and expose the underlying carotid bifurcation
              -	Identify and protect Hypoglossal nerve (CN XII) 
              -	Ansa cervicalis may be ligated if necessary for exposure
        -	Dissect carotid bifurcation 
              -	Careful to not stimulate carotid body and cause hemodynamic instability
        -	If access to distal internal carotid artery is required, may consider subluxation of mandible or mandibular osteotomy to gain appropriate exposure

Vertebral Artery
-	Anatomy 
        -	First cephalad branch of subclavian artery 
        -	Divided into 3 parts
              -	Part I (proximal)
                    -	Origin at subclavian artery to C6 (enters transverse foramen)
                    -	External landmarks: found between two heads of SCM (sternal and clavicular) and clavicle
                    -	Vertebral artery runs between anterior scalene and longus colli muscles
              -	Part II 
                    -	Courses through bony vertebral canal (transverse foramen) from C6 to C1
              -	Part III (distal)
                    -	Courses outside vertebral canal from C1 to base of skull (enters through foramen magnum) and joins contralateral vertebral artery to form basilar artery (part of circle of Willis)
-	Exposure
      -	Will focus on exposure of Proximal Vertebral Artery
      -	Supraclavicular transverse incision extending between the sternal and clavicular heads of the sternocleidomastoid (SCM) muscle
      -	Incise platysma and continue dissection into base of triangle
      -	Carotid sheath is first vascular structure identified
              -	Jugular vein  lateral
              -	Common carotid  medial
              -	Vagus nerve  posterior 
      -	Identify scalene fat pad between two heads of SCM and clavicle, dissect to expose anterior scalene muscle
              -	Careful of phrenic nerve as it runs on surface of anterior scalene muscle 
        -	Divide anterior scalene muscle to visualize the subclavian artery and two of its branches (thyrocervical trunk and internal mammary artery). Control each branch. 
        -	Proximal VA is located deep to supraclavicular artery and anterior to groove between C7 vertebral body and transverse process 
              -	Proximal VA is between anterior scalene (laterally) and longus colli (medially)
              -	Best found through palpation with tip of index finger within groove of C7 vertebral body and transverse process	
              - Avoid injury to vertebral venous plexus as it is anterior to VA
        -	Use right angle clamp to dissect out VA	

Detailed exposure techniques can be found here: **(INSERT ZOTERO REFERENCE KWON ET AL)**  


**Tenets of Cerebrovascular Injury** 

-	Control of catastrophic bleeding --- with manual compression and resuscitation
-	Ensure that injury is not causing airway compromise 
-	Evaluate the neurological status of patient --- have a thorough baseline neuro exam and look for signs of ischemia, hemispheric stroke, arousability/mental status
-	Prevent secondary injury --- maintain blood pressure to maintain cerebral perfusion pressure and prevent hypoxia 
-	After appropriate exposure and control, recommend that patient be heparinized during carotid artery repair. Be prepared for bleeding, especially in patients with polytrauma. Resuscitate appropriately until repair completed. 

**Blunt Carotid Injury Patterns**

-	Mechanism of Injury: history of a motor vehicle collision, blunt force to neck, severe hyperextension/rotation/flexion injuries, etc. 
-	External signs of injury: marks on neck or around the thoracic inlet (ie seatbelt sign), unilateral neurodeficits (suggestive of hemispheric ischemia or stroke)
-	Accompanying injuries: concomitant closed head injuries (TBIs), complex facial fractures, cervical vertebral body and transverse foramen fractures


**Blunt Carotid Injury Classification System and Management** 

Diagnostic Imaging: 
-	CTA is gold standard. Limitation, image is static. 
-	US duplex combines B mode ultrasound with pulse doppler. Limitation in visualization of injury above the angle of the mandible (distal internal carotid and proximal thoracic outlet)
-	These imaging modalities are used to assess visibility of defect (ie dissection, thrombus, intimal flap) and degree of flow limitation by measured velocity. 
 
Denver Grading Scale for Blunt Vascular Carotid Injury (BCVI) **(INSERT ZOTERO REFERENCE: BIFFL ET AL)**

-	Grade I: Luminal irregularity or dissection with < 25% luminal narrowing 
-	Grade II: Dissection or intraluminal hematoma with  25% luminal narrowing, intraluminal thrombus, or raised intimal flap 
-	Grade III: Pseudoaneurysm 
-	Grade IV:  Occlusion
-	Grade V: Transection with Free Extravasation 

Treatment Options 
-	Considerations 
        -	Consider the overall injury severity of the trauma patient with any concomitant injuries that may prevent antiplatelet/anticoagulation initiation (ie TBI) or intervention
              -	Oftentimes have other blunt vascular injuries, such as aortic injuries, that need to be addressed as well
              -	Can consider delay of intervention (approx. 7 days) to decrease risk of neurological events
        -	If there is a change in the neurological exam after initiation of medical management, considered as failure of medical therapy and surgical intervention is indicated
        -	If unable to obtain a reliable neurological exam, can consider transcranial doppler or EEG if accessible, but is not mandatory
        -	Continue supportive measures to prevent secondary injuries, ie optimize cerebral perfusion pressure 
        -	Repeat imaging with change in neurological status
        -	Serial imaging important in assessing whether injury has evolved
        -	Endovascular repair usually involves transfemoral approach for access with arch aortogram, selection of defective carotid vessel, and placement of the endovascular treatment (ie: stent graft, bare metal stent, coil embolization of pseudoaneurysm)

-	Carotid Artery Injury Treatments 
        -	Grade I and II injuries are recommended to be treated medically with antiplatelet/anticoagulation therapy if patient is able to tolerate 
              -	Repeat CTA imaging in 5-7 days while inpatient to ensure that injury has not evolved in acute phase
              -	Follow-up CTA imaging within 1 month of discharge as outpatient to assess healing. Imaging needs to be continued until there is resolution of lesion. 
              -	Grade I lesions typically improve over time, Grade II lesions or higher have a greater chance of evolving (approximately 60%)
        -	Accessible Grade III and V carotid injuries should be repaired. 
              -	Grade III pseudoaneurysms should be repaired endovascularly with a covered stent (ie stent graft that covers the opening of the pseudoaneurysm) or bare metal stent (maintains integrity of the wall). Through these stents, coils can be placed to thrombose the pseudoaneurysm. Flow needs to be maintained. 
              -	Open approach can be considered for Zone II injuries 
        -	Grade I and IV carotid injuries, and inaccessible Grade II and III carotid injuries should be treated with medical management (antiplatelet/anticoagulation)
              -	Grade IV injuries without associated injuries that preclude use of anticoagulation, would heparinize and proceed with anticoagulation (avoid bolus of heparin, goal PTT 50-70 within first 24 hours). Monitor for bleeding. Transition to oral anticoagulant for 30-90 days. 
              -	Grade IV injuries with contraindications for anticoagulation, would recommend dual antiplatelet therapy (aspirin and clopidogrel)
              -	Overall, anticoagulation is preferred if there are no contraindications. If there are, then dual antiplatelet therapy is preferred.  
        -	Grade V and persistent Grade III should be repaired surgically 

-	Vertebral Artery Injury Treatments 
        -	Special Considerations
              -	Attempt to determine whether the dominant or diminutive vertebral artery is involved
              -	Treatment is again based on severity with grading system (Denver Grading Scale), consider location and extent 
              -	Consider any associations with a basilar artery defect or posterior circulation stroke
        -	Grade I-IV vertebral injuries should be treated with medical management as first line, approximately 90% self-resolve
              -	Endovascular interventions can be considered for those patients with severe injury and/or who are symptomatic
                      -	Usually performed by neuro-interventional as the injury is located within the skull base
              -	Serial imaging still needs to be performed to evaluate healing (similar manner to carotid injury) 
              -	In the setting of trauma, if there is an expanding hematoma due to vertebral artery injury and open approach is performed to control bleeding, vertebral artery is usually ligated in this scenario
                      -	Vertebral reconstructions are not well-suited for traumas

Guidelines and treatment algorithm can be found here: ***(INSERT ZOTERO GEDDES ET AL)*** 

**Overview of Neck Zones for Penetrating Trauma** 
Levels
-	Zone I: thoracic inlet (sternal notch) to level of cricothyroid cartilage
-	Zone II: cricothyroid cartilage to angle of mandible
-	Zone III: angle of the mandible to the base of the skull

Image of neck zones can be found here: < https://aneskey.com/penetrating-neck-trauma-2/>

Management 

-	Injury to Zone I and Zone III 
        -	If hemodynamically stable, requires imaging prior to consideration of intervention/repair 
              -	Imaging: CTA and/or angiogram
              -	CTA positive >> Endovascular intervention vs. OR 
              -	Consider concomitant injury to aerodigestive tract as well
        -	If hemodynamically unstable with/without hard signs (aerodigestive or neurovascular injuries)
              -	Tamponade >> Secure Airway for Air leak/hematoma >> OR
              
-	Injury to Zone II (violation of platysma) 
        -	Direct to OR for exploration, especially with high suspicion of carotid injury and neurovascular deficits (hard sign)
        -	Repair carotid artery if injured, even if thrombosed
              -	The risk of reperfusion injury causing an intracranial hemorrhage was traditionally considered a reason to not repair. 
              -	However, overall mortality and final neurostatus has shown to be better with intervention, even if patient started with neurodeficits.
              -	Requires communication with operative team (ie anesthesia) to mitigate hypoperfusion
        -	Consider concomitant injury to aerodigestive tract
        -	Scenario: Patient has large cerebral infarct on same side as carotid injury. 
              -	Continue to OR to explore
              -	Preferred management is to repair and attempt reperfusion of hemisphere, possibly perfusing a previously ischemic penumbra
              -	Ligation usually not an option.	
                    -	Unless carotid is completely occluded and there is no retrograde bleeding after attempted thrombectomy >> there is potential for ligation, but not readily considered. 
              -	Observation without repair can risk evolvement of injury into pseudoaneurysm with need to repair and reoperate days/weeks later
        -	Outcome of this repair is likely a stroke
        
-	Repair Techniques
        -	Appropriate exposure of carotid artery (see above for details) 
        -	Gain proximal and distal control 
        -	Identify injury
        -	Remove burden of thrombus (if present)
              -	Thrombus burden must be removed prior to repair
              -	Via flushing from back-bleeding and forward-bleeding of internal carotid artery
              -	Via Fogarty or thrombectomy catheter, carefully 
                    -	Recommend 2 French or 3 French size 
                    -	Pass only 1-2 cm up into distal internal carotid
        -	Debride any devitalized tissue prior to repair
              -	May cause defect to be larger, but repair requires healthy tissue planes
        -	Repair carotid 
              -	Heparinization required when repairing carotid artery and/or clamping
                    -	Be efficient, as these patients usually have significant polytrauma and other areas that are susceptible to bleeding which will need to be controlled when heparin administered
              -	Repair options 
                    -	Primary Repair 
                    -	Patch Repair 
                          -	Used for short-segment injuries
                          -	Conduits
                                  -	Dacron or bovine pericardium 
                                  -	Autologous vein
                                            -	Not first-line due to extra time required for harvest
                                            -	Used if there is concern for contamination (ie shot gun wound with large soft tissue injury, aerodigestive injury, etc)
                    -	Interposition Graft
                          -	Used for long-segment injuries 
                          -	Conduits: 
                                  -	Autologous vein (great saphenous vein) 
                                            -	Good option when there is concomitant aerodigestive injury 
                                            -	Consider shunting while waiting for harvest 
                                  -	Synthetic (ie Dacron/PTFE/Goretex)
                    -	Transposition of Internal Carotid Artery to External Carotid artery 
                          -	Can avoid need for shunting 
                          
-	Consideration of Shunting 
        -	In most cases, shunting not required because adds extra level of complexity to trauma case that may not have the luxury of time
        -	Cases where shunting can be used:
              -	Interposition graft repair for long-segment carotid injury that requires vein harvest > shunting can be performed to reperfuse the affected brain hemisphere while waiting for conduit preparation
        -	Typical shunts used: Argyle or Sundt
        
-	Assessment of Repair 
        -	Assess repair with removal of thrombus and once back-bleeding restored
        -	Duplex US (continuous wave doppler)
              -	Best option when time is limited to assess flow 
              -	Usually also used most often with elective carotid repairs
        -	Intraoperative Arteriograms 
              -	If have luxury of time to perform and set up is appropriate, used for completion studies
              -	Documents complete evacuation of distal thrombus 
              -	Performed by placing 18-gauge butterfly needle in CCA (remove air) and inject contrast to view repair and distal internal carotid to ensure lack of thrombus
              
-	Consideration of Drains 
        -	If there is contamination (aerodigestive injury), then can place a small JP drain (7mm) to protect the repair
        -	Buttressing repair in contaminated cases usually not used as the neck is highly vascular
              -	Can consider a muscle flap to separate arterial injury from esophageal injury (commonly not seen) with gross contamination

Guideline Algorithm for Penetrating Neck Trauma: ***(INSERT ZOTERO REFERENCE SPERRY ET AL)*** 

Paper that first presented the anatomic neck zones: ***(INSERT ZOTERO REFERENCE ROON ET AL)***


**Internal Jugular Vein Injuries** 

-	If patient has significant TBI, consider repairing Internal Jugular venous injuries to maintain venous outflow to that side of the brain --- especially if there is a transduced venous pressure gradient. Cases include patients who have intracranial hypertension and require a decompressive craniectomy. 

-	Otherwise, able to ligate without much consequential effects in most patients due to appropriate amounts of collateral outflow. 



EAST Blunt Cerebrovascular Injury Guidelines

<https://www.east.org/education/practice-management-guidelines/blunt-cerebrovascular-injury>

Check out Rich's Vascular Trauma

<https://www.elsevier.com/books/richs-vascular-trauma/9781455712618>



## Endovascular Approaches

**09 Feb 2021**: *Kevin Kniery, MD, MPH; Marlin "Wayne" Causey MD; Todd
Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Rasmussen_Endovascular_Trauma.mp3")

```

**Blunt Thoracic Aortic Injury (BTAI)**

Epidemiology
	- Rare, but lethal 
	- Less than 1% of all blunt traumas
	- Second leading cause of death in blunt trauma
	- Due to high velocity mechanism of injury, usually are polytrauma patients with other severe injuries (ie intracranial injury) 
	- Common mechanisms 
	    - High-speed motor vehicle collision 
	    - Motorcycle collisions
	    - Pedestrian vs. motor vehicle 
	    - Falls from heights 
	- Most patients have unsurvivable devastating injuries (75%) and pass prior to arrival to hospital for definitive care
	- Rates of paraplegia, rates of blood utilization, and mortality rates have decreased with prominence of endovascular repair of injuries

Anatomy 
	- Aortic arch >>  relatively mobile
	    - Not common location for blunt injury
	- Descending Thoracic aorta 
	    - Tethered 
	    - Junction just distal to left subclavian artery is where most injuries occur (within 0.5-2cm)
	    - Tear injury due to deceleration 

Diagnostic Modalities 
	- Be aware that patients likely have other injuries 
	- Recommend CTA chest, abdomen, pelvis 
	- Perform CTA follow-through to upper/lower extremities if those areas are of concern for injury

Grading Systems
 
	- SVS BTAI Grading System (SVS guidelines ***INSERT ZOTERO REFERENCE LEE ET AL***) 
	    - Grade I: Intimal Tear 
	    - Grade II: Intramural Hematoma 
	    - Grade III: Pseudoaneurysm
	    - Grade IV: Rupture

	- Harborview Grading System (**INSERT ZOTERO REFERENCE HENEGHAN ET AL and STARNES ET AL***) 
	    - Classification scheme helps dictate which patients can be managed operatively vs. nonoperatively 

	    - Minimal Injury
	          - Absence of aortic external contour abnormality
	          - Intimal tear and/or thrombus of <10mm in length or width 
	          - Treatment: No intervention
                - Surveillance with follow-up imaging 
	              - Antiplatelet therapy
	              - CTA at least within 30 days of injury
	    - Moderate Injury
	          - External contour abnormality, includes pseudoaneurysms
	          - Intimal tear  10mm in length or width 
	          - Treatment: Delayed repair, semi-elective
	               - Stabilization of concomitant injuries 
	               - Impulse control 
	                      - Short-acting beta blocker (ie esmolol) 
                        - SBP < 120 (coordinate with neurosurgery for TBI patients) 
	               - Repair within 1 week (TEVAR)
	                      - For those patients with severe TBI who are unable to undergo anti-impulse therapy due to maintenance of cerebral perfusion pressure, will recommend repair sooner
	          - Repeat CTA chest within 5-7 days for follow-up
	     - Severe Injury
	           - Active extravasation (free contrast extravasation or hemothorax at thoracotomy)
	           - Hemodynamic instability
	           - Left subclavian artery (LSA) hematoma > 15 mm 
	           - Treatment: Emergent repair
	               - BTAI takes first priority 

SVS Guidelines for Blunt Thoracic Aortic Injury can be seen here: <https://www.jvascsurg.org/article/S0741-5214(10)01924-5/fulltext> 


**Endovascular Treatment Options for Blunt Thoracic Aortic Injury**

Overview 
	- Endovascular repair of BTAI has better outcomes than Open repair
	- Need to consider that most TEVAR devices are designed for aneurysmal disease
	        - Generally require a larger diameter and their compliance is not ideal for these smaller/normal caliber aortic arche sizes 
	        - Average size of aorta proximal to injury site is 19mm (smallest devices are at 22mm)
	        - Previously, iliac devices used to treat aortic injuries 
	- There are new advanced TEVAR devices that are available for non-aneurysmal disease with normal caliber aortic sizes. 
	        - ex: GORE TAG Conformable Thoracic Stent Graft with ACTIVE CONTROL System
	- Recommend heparinization if not contraindicated
	        - Ie: TBI, femoral shaft fracture with hematoma, solid organ injury, etc. 
	        - TEVAR can still be performed safely without heparin
	        - Higher threshold for heparinization in COVID+ patients due to their hypercoagulable state
	        - Goal Activated Clotting Time (ACT)  > 250 
	              - Ensure that ACT machine is available prior to start of case

Methods of Endograft Size Measurement
  - CTA imaging
        - Used to help measure aortic size
        - Need to consider that patient is usually hypovolemic and constricted --- true aortic size may be under-represented on imaging when choosing the appropriate device for repair
        - Recommend over-sizing stent graft by 20-25% from original measurement 
  - Intravascular Ultrasound (IVUS)
        - During angiogram, can confirm or redirect sizing in real-time

Access
  - If iliac size is >7-8 mm, can use transfemoral access
  - Small iliacs (< 6mm) may need conduits
        - Recommend retroperitoneal approach for exposing common femoral and/or external iliac and using a dacron conduit (**INSERT ZOTERO REFERENCE CRIADO ET AL***)
  - If iliacs are diseased (ie calcium burden), use of conduits would be difficult
        - Can consider putting mineral oil over sheath to help pass and deliver the endograft
  - Femoral cutdowns considered to ensure safe delivery and removal of device

Positioning of Device
  - Most important aspect is to cover the proximal defect and seal the entry point
  - Average distance between the injury and the LSA is approximately 5.8mm (located at aortic isthmus)
  - Recommend 2cm coverage proximally of healthy vessel and 2cm coverage distally
        - With these recommendations, approximately 40% of TEVAR repairs cover the LSA
        - Pre-emptive revascularization of LSA is not necessarily required, most patients do fine without arm claudication
                - Complications to consider when covering LSA
                    - Stroke
                           - If there is concern for vertebral perfusion, LSA must be revascularized prior to covering
                                - left vertebral artery may be dominant or significant disease in right vertebral artery
                    - Arm claudication and ischemia
                    - Spinal cord ischemia 
                           - Rare
                           - Collateral pathways between LSA and IMA (internal mammary artery) can be compromised
                           - Spinal cord ischemia can present at end of case or delayed 24-72 hours post-operatively 
                           - Spinal cord drainage not routinely done to prevent paraplegia unless there is concern if abnormal neuro-exam.
                                - If abnormal neuroexam at end of case, elevate blood pressure to improve perfusion and consider placing a drain at that time
                                 - There is high risk of placing and monitoring drains if not needed
              -	Relative contraindications to covering LSA without revascularization 
                    -	Dominant Left Vertebral Artery perfusion to brain
                    -	Right Vertebral Artery atherosclerotic disease
                    -	Hemodialysis access on ipsilateral arm 
                    -	Prior CABG with LIMA (left internal mammary) to LAD (left anterior descending artery) 
  -	Some devices (ie GORE conformable as above) allow control of first stent ring to better conform/lay down the stent within the arch

Malpositioning of Devices
  -	Can deploy too proximal or too distal 
  -	Bird Beaking
        -	Where the lesser curve of the graft is not well approximated to the lesser curve of the aorta (ie the lifting of the inferior aspect of the most proximal endograft with the first covered stent ring now lifted off into the inner curve of the aortic arch) (**INSERT ZOTERO REFERENCE FROHLICH ET AL**)
        -	Can cause endoleaks
        -	Methods to avoid bird beaking
                -	Use stiff wire to get into the aortic root (ie Amplatz or Lunderquist wires) with forward pressure to have a constant tension 
                      -	Wire should be snug against the greater curvature of the aorta
                -	Can consider taking the graft over the origin of the LSA
                      -	Allows more room in the flatter inner arch to lay the graft down

Timing of Repair 
  -	SVS Grade 3 or Harborview Classification of Moderate Injury 
        -	Delayed repair 24-48 hours after injury is acceptable and there is a survivable benefit (***INSERT ZOTERO REFERENCE HERE ALARHAYEM ET AL***)
        -	Some delays can be up to 5-7 days with medical management
        -	Allows time for operative planning and optimization with anti-impulse therapy/resuscitation
        -	Also enables evaluation of any underlying infectious process that may compromise and infect the endograft
  -	SVS Grade 4 or Evolving SVS Grade 3/Harborview Moderate injuries 
        -	More urgent repair (within 24 hours) 


**Blunt Abdominal Aortic Injuries and Endovascular Management**

Epidemiology 
  -	Very Rare
  -	Commonly associated with polytrauma patients, especially with concomitant blunt thoracic aortic injuries 
  -	Similar mechanisms of injury as BTAI 

Anatomy 
  -	Zone 1: Diaphragm to SMA (superior mesenteric artery)
  -	Zone 2: SMA to renal arteries
  -	Zone 3: Inferior renal artery to aortic bifurcation

Overview 
  -	Zone 1 and 3 can be repaired endovascularly (above the SMA and below the renal arteries) 
  -	Can cover the celiac artery safely if there is no obvious compromise of the other mesenteric arteries (ie: inferior mesenteric artery, internal iliac arteries, prior gastrectomy or other complex intra-abdominal operation)
  -	Zone 2 injuries typically repaired with open approach
  -	Endovascular approach can minimize risk of contamination if there is bowel spillage etc. 
  -	Want to use minimum amount of endograft that will appropriately cover injury 
	2cm coverage proximally and 2 cm coverage distally 

Diagnostic Imaging 
  -	CTA 
  -	IVUS 
  -	Also used for operative planning and evaluating extent of defects into branches of the mesentery

Types of Endografts Used 
  -	TEVARs 
  -	Covered stents 
  -	Aortic cuffs 	

  - Can use combination of all grafts types to get the appropriate coverage


**Blunt Axillo-Subclavian Injuries and Endovascular Management** 

Epidemiology 
  -	Very rare
  -	Low incidence due to protection of vessels from surrounding bony structures and high degree of exsanguination with devastating injuries
  -	Endovascular repair preferred due to location of these injuries
  -	difficult to access via open methods with proximal/distal control due to protective anatomic nature of overlying skeleton

Anatomy 
  -	Thoracic Outlet 
        -	Axillo-subclavian vein located anterior to anterior scalene and runs between the first rib and clavicle through the subclavius muscle/tendon
        -	Axillo-subclavian artery is located in the arterial triangle
                -	More lateral than the axillo-subclavian vein 
                -	Courses posteriorly to anterior scalene

Access 
-	Transbrachial or transradial access preferred 
-	Can combine transbrachial and transfemoral 
        -	Provides more options for delivering stent grafts with difficult aortic arches (anterograde and retrograde approach) via a through wire
-	With transbrachial access, brachial cutdown can be more reliable in trauma setting
-	Can also consider HYBRID approach 
        -	Gain proximal and distal control via through wires from transbrachial/transradial and then perform the open repair
        -	Can perform coil embolization of subclavian arteries to gain control 
                -	Caution to vertebral perfusion
        -	Stent graft to temporize injury and maintain in-line flow, then have interval axillary artery repair with a bypass (within a week to a month later) 

Types of Endografts used 
-	Self-expanding stents (ie Viabahn) 
        -	Traditional approach 
        -	May provide greater flexibility
        -	Helpful for more distal injuries
-	Balloon-expandable stents
        -	Newer devices, also flexible
        -	Has more precise landing (especially if injury is near branch points)





Seminal Papers in Blunt Thoracic Aortic Injury

-   AAST 1997 Paper: <https://pubmed.ncbi.nlm.nih.gov/9095103/>

-   AAST 2008 Paper: <https://pubmed.ncbi.nlm.nih.gov/18545103/>

-   JVS 2011 Paper: <https://pubmed.ncbi.nlm.nih.gov/20974523/>

-   Timing of repair of BTAI JVS
    2020: [https://www.jvascsurg.org/article/S0741-5214(20)31575-5/fulltext](https://www.jvascsurg.org/article/S0741-5214(20)31575-5/fulltext){.uri}

Dr. Ben Starnes' podcast on Behind The Knife on BTAI: <https://bit.ly/2LuycWq>




